---
title: "Application Data"
vignette: >
  %\VignetteIndexEntry{Application Data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(STOPeData)
```


The STOPeData app uses what the developers of `Golem` call the ["stratégie du petit r"](https://engineering-shiny.org/structuring-project.html?q=document#communication-between-modules); that is, almost all the app's data is stored in a big reactiveValues object in the main server function.

```{r}
#| label: app_data

initialise_userData <- function() {
  list(
    ENTERED_BY = character(0),

    # Standard validated data ----
    # All userData and module_state$data data is stored in a tabular (tibble) format centrally, even for campaign and reference (which currently only have one row)
    # This means we can use a consistent set of functions to check for presence (nrow(tibble) > 0), and not have any nasty surprises when we expect one and get the other
    # Data entry modules
    sitesData = initialise_sites_tibble(),
    sitesDataValid = FALSE,
    parametersData = initialise_parameters_tibble(),
    parametersDataValid = FALSE,
    compartmentsData = initialise_compartments_tibble(),
    compartmentsDataValid = FALSE,
    referenceData = initialise_references_tibble(),
    referenceDataValid = FALSE,
    campaignData = initialise_campaign_tibble(),
    campaignDataValid = FALSE,
    methodsData = initialise_methods_tibble(),
    methodsDataValid = FALSE,
    samplesData = initialise_samples_tibble(),
    samplesDataValid = FALSE,
    biotaData = initialise_biota_tibble(),
    biotaDataValid = FALSE,
    samplesDataWithBiota = tibble(NULL),
    measurementsData = initialise_measurements_tibble(),
    measurementsDataValid = FALSE,

    # CREED Data
    datasetDetails = tibble(NULL),
    creedRelevance = initialise_CREED_data_tibble(),
    creedReliability = initialise_CREED_data_tibble(),
    creedScores = initialise_CREED_scores_tibble(),
    creedReport = "",

    # CREED reactive objects that just exist to trigger reactivity. Probably bad coding!
    creedGetData = 0, # watched by multiple observers in nested CREED modules. +1 every time we input$get_data in mod_CREED
    creedCalculateScores = 0, # same

    # LLM extracted data and metadata ----
    schemaLLM = "",
    promptLLM = "",
    rawLLM = "",
    pdfPath = NULL,
    campaignDataLLM = tibble(NULL),
    referenceDataLLM = tibble(NULL),
    sitesDataLLM = tibble(NULL),
    parametersDataLLM = tibble(NULL),
    compartmentsDataLLM = tibble(NULL),
    methodsDataLLM = tibble(NULL),
    samplesDataLLM = tibble(NULL),
    biotaDataLLM = tibble(NULL),
    samplesDataLLM = tibble(NULL),

    # LLM extraction status flags ----
    llmExtractionComplete = FALSE, # tracks if the LLM data extraction process has completed, or the user has pressed the dummy data button
    llmExtractionSuccessful = FALSE, # tracks if the LLM data extraction process (or dummy data) returned a tibble in the expected format
    llmPopulateModules = FALSE, # tracks if the user has sent LLM data to modukles

    llmExtractionComments = tibble(NULL),

    # Import data from save status flags ----
    saveExtractionComplete = FALSE,
    saveExtractionSuccessful = FALSE
  )
}
```

This has its pros and cons.

- Pro: No need to write specific code to for each transfer of data between modules
- Pro: Data can be initialised using a list function, as above. makes testing easier since we can mock the app data structure outside of a shiny session
- Pro: Better centralised overview of flags, etc.
- Con: Assigning to a new value in a list creates it, so there's no way to defend against e.g. typos in flags, which will create a new flag (that doesn't do anything)
- Con: At least the way I'm using it, leads to long, difficult-to-read variable names like `session$userData$reactiveValues$sitesData`
- Con: Our checks for data presence are fairly messy, because we have to use `!is.null(data) && nrow(data) != 0` for tibbles, but different checks for strings, vectors, etc. 

It would probably be much safer to use an R6 object because then we could defend against invalid inputs. And I have actually done this in another branch, but it's not up to date with other changes, and I think it's something I'll save for when I (perhaps) have more time and money.