---
title: "eData Application"
subtitle: "How the application runs"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(STOPeData)
```

This is how we run the application on our server (see `Dockerfile`):

```{r}
#| eval: false
shiny::runApp('app.R', host='0.0.0.0', port=3838)

# this is how I run the app locally when I'm developing it
golem::run_dev() 
```

It calls `~app.R`. This sets up the `logger` framework, sets some Golem options, and calls `STOPeData::run_app()`, located in `~.R`.

```{r}
#| eval: false
#| label: app.R
logger::log_appender(logger::appender_stdout)
logger::log_layout(logger::layout_json())
logger::log_threshold(logger::INFO)

logger::log_messages()
logger::log_warnings()

pkgload::load_all( # without pkgload dependencies are loaded very slowly
  export_all = FALSE,
  helpers = FALSE,
  attach_testthat = FALSE,
  quiet = TRUE
)
options(shiny.maxRequestSize = 20 * 1024^2) # increase permitted file upload size
options(golem.app.prod = FALSE)
options(bslib.color_contrast_warnings = FALSE)
STOPeData::run_app() # add parameters here (if any)

```

`STOPeData::run_app()` calls `~/R/run_app.R`, which is a wrapper function created by the `Golem` package.

```{r}
#| eval: false
#| label: STOPeData::run_app()

#' Run the Shiny Application
#'
#' @param ... arguments to pass to golem_opts.
#' See `?golem::get_golem_options` for more details.
#' @inheritParams shiny::shinyApp
#'
#' @export
#' @importFrom shiny shinyApp
#' @importFrom golem with_golem_options
#' @importFrom logger log_appender log_layout log_threshold log_messages log_warnings layout_json

run_app <- function(
  onStart = NULL,
  options = list(),
  enableBookmarking = "server",
  uiPattern = "/",
  ...
) {
  # set up logger logging for gcp

  with_golem_options(
    app = shinyApp(
      ui = app_ui,
      server = app_server,
      onStart = onStart,
      options = options,
      enableBookmarking = enableBookmarking,
      uiPattern = uiPattern
    ),
    golem_opts = list(...)
  )
}

```

This calls `shiny::shinyApp()` on the ui and server objects created by `~/R/app_ui.R` and `~R/app_server.R`.

```{r}
#| eval: false
#| label: ui and server functions

app_ui <- function(request) {
 # ui goes here
}

app_server <- function(input, output, session) {
 # logic goes here
}
```

In actual fact, though, the application is heavily modularised/functionalised. This means that the functional code you're looking for is more likely to be in the module files (`~/R/mod_*.R`) or function files (`~/R/mod_*_fct.R`,`fct_*.R`, and `utils_*.R`).
