---
title: "LLM Evaluation"
subtitle: "Comparing the quality, speed, and cost of LLM extraction"
knitr:
  opts_chunk:
    collapse: true
execute:
  error: true
format: 
  html:
    code-fold: true
  pdf: 
    echo: false
    warning: false
---

```{r}
#| label: setup
#| echo: false
#| include: false

library(STOPeData)
library(ellmer)
library(tidyverse, mask.ok = c("filter", "lag"))
library(flextable)
library(arrow)
library(sf)
library(eDataDRF)
library(pointblank)
library(irr)
library(here)


devtools::load_all() # remove once STOPeData is rebuilt

set_flextable_defaults(
  font.size = 10,
  # font.family = 'Aptos',
  digits = 2,
  theme_fun = theme_vanilla,
  padding = 1
)

# Create a location to save LLM API calls, if it doesn't exist.
LLM_CACHE_DIR <- here("data-raw/llm_cache/calls")
dir.create(LLM_CACHE_DIR, recursive = TRUE, showWarnings = FALSE)

i_am("README.md")
```

# Introduction

The `ellmer` package includes functions that connect to a wide variety of local and remote APIs from a variety of providers. Put simply, we have a great many options. Which are better, and which are worse? We might assume that more expensive models are more effective at data extraction, and that larger and more complex files are more difficult - and require more tokens - than simpler models when extracting data. But can we prove any of these assumptions?

Token size defines the memory space available to the LLM, and the cost of extraction. 

# Methods

```{mermaid}
%%| label: workflow-diagram
%%| fig-cap: "Overview of the LLM evaluation pipeline."
flowchart TD
    pdfs["PDFs tibble<br>4 papers"]         --> grid
    models["Models tibble<br>3 providers"]  --> grid
    tok["Token limits<br>3k / 6k / 9k<br>n replicates"] --> grid

    grid["Experiment grid<br>run_extraction_batch"]  --> cache

    cache{"RDS cache hit?"}
    cache -- Yes --> load["Load cached RDS"]
    cache -- No  --> api["LLM API call<br>extract_pdf_with_llm"]

    api -- success --> save["Save RDS"]
    api -- failure --> err["Record error<br>HTTP code + message"]

    save  --> results
    load  --> results
    err   --> results

    results["calls_results<br>call list-column"]

    results --> flat["rebuild_flat_parquet<br>Successful calls only"]
    flat    --> parquet[("calls_flat parquet<br>long / flat format")]

    results --> report["batch_run_report<br>success · error · tokens<br>by pdf · model · limit"]

    parquet --> status["call_status<br>success · empty-table<br>effectiveness flags"]
    parquet --> validate["validate_all_calls<br>pointblank agents<br>per table"]
    parquet --> canon["Canonical comparison<br>vs. human-extracted data<br>controlled + continuous vars"]
    parquet --> kappa["Kappa statistics<br>inter-rater reliability<br>within- and cross-model"]
```

## Study Materials

We selected four studies [@tbl-pdfs] that were previously identified as high-quality (score >= 4), covering a variety of time periods, formats, topics, and methods of presenting data.  

```{r}
#| label: tbl-pdfs
#| tbl-cap: "Study materials (portable document format files) selected for LLM evaluation. OCR (Optical Character Recognition) refers to cases where the paper used has been scanned from a printed copy and automatically converted to text. Depending on the quality of this process, information may be lost or corrupted; we would expect such files to be more difficult to extract data from."

pdfs <- tibble(
  id = c("a", "b", "c", "d"),
  path = c(
    here("data-raw/1997RitterhoffTraceMetalsIn_MS.pdf"),
    here("data-raw/2002HellandTransportAndSedimentation_MS.pdf"),
    here("data-raw/2021WojtasikTemporalIsolationBetween_MS.pdf"),
    here("data-raw/2022KohlerDistributionPatternOf_MS.pdf")
  ),
  title = c(
    "Trace metals in field samples of zooplankton from the Fram Strait and the Greenland Sea",
    "Temporal isolation between two strongly differentiated stocks of the Greenland halibut (<i>Reinhardtius hippoglossoides</i> Walbaum, 1792) from the Western Barents Sea",
    "Transport and sedimentation of Cu in a microtidal estuary, SE Norway",
    "Distribution pattern of mercury in northern Barents Sea and Eurasian Basin surface sediment"
  ),
  year = c(1997L, 2002L, 2021L, 2022L),
  comment = c(
    "Well-OCR'd paper with summary statistics of trace metal concentrations reported across wide range zooplankton taxa. Sampling sites reported in low-quality map. Complex, uneven data structure due to random nature of zooplankton sampling. Methodology reported in moderate detail.",
    "Native PDF.  Sites reported in map. Cu concentrations (and flux) reported from a variety of sediment traps at different timepoints. Methodology reported in moderate detail.",
    "Native PDF. Study primarily concerns evolution but includes heavy metal analysis. Sites reported as linear trawls with start/end coordinates. Raw Cu concentrations given for each individual, reported as table. Methodology reported in high detail.",
    "Native PDF. Sampling sites and Cu concentrations only reported as summaries/graphics in paper, but raw data (Cu concentrations, site coordinates, depth) provided on request and prepended to file."
  ),
  # Fill in after reviewing each PDF
  num_images = c(1, 4, 8, 5),
  num_tables = c(7, 2, 8, 2)
) |>
  mutate(
    info = map(path, pdftools::pdf_info),
    file_size_kb = map_dbl(path, \(p) file.size(p) / 1024),
    num_pages = map_int(info, \(x) x$pages)
  ) |>
  select(-info)

pdfs |>
  select(-path) |>
  flextable() |>
  autofit() |>
  colformat_double(j = c("file_size_kb")) |>
  colformat_int(j = c("year"), big.mark = "") |>
  set_header_labels(
    id = "ID",
    title = "Title",
    year = "Year",
    comment = "Comment",
    num_images = "n Images",
    num_tables = "n Tables",
    file_size_kb = "File Size (kb)",
    num_pages = "n Pages"
  )
```

## Large Language Models

The R `ellmer` package includes API integrations for a variety of Large Language Models from various providers. We selected general, "middle-of-the-road" multimodal models from three of the most prominent providers - Anthropic, Google Gemini, and OpenAI, as well as immediately higher and lower tier models. In addition, we wished to validate the performance of local and/or open-source LLMs. This proved technically difficult, and with the computational resources available we were unable to run models locally at acceptable levels of performance (models could technically be run, but lacked the available memory to load PDF files). As a compromise, we also include the open model Gemma. Gemma is developed by Google (and thus accessible via their API to be run on their servers), but can also be loaded and run locally using services such as Ollama (which supports structured data extraction) and LMStudio and accessed locally with ellmer via their APIs. 

Our selected models are summarised in [@tbl-models]. Details are sourced from providers' model overview pages.

Sources for model comparisons: 
https://platform.claude.com/docs/en/about-claude/models/overview, 
https://developers.openai.com/api/docs/models/compare, 
https://ai.google.dev/gemini-api/docs/models, 
https://ai.google.dev/gemini-api/docs/pricing


```{r}
#| label: tbl-models
#| tbl-cap: "Summary of LLM models included in the evaluation."

# TODO: Fill out list

models_tbl <- tribble(
  ~provider                          , ~model_label                         , ~model_id                                                                                                                                                                     ,
  ~context_window_k                  , ~max_output_k                        ,
  ~cost_input_mtok                   , ~cost_output_mtok                    ,
  ~knowledge_cutoff                  , ~training_cutoff                     , ~description                                                                                                                                                                  ,
  # ---- Anthropic ---------------------------------------------------------------
  "Anthropic"                        , "Claude Haiku 4.5"                   , "claude-haiku-4-5"                                                                                                                                                            ,
  "200k"                             , "64k"                                ,
  "$1"                               , "$5"                                 ,
  "Feb 2025"                         , "Jul 2025"                           , "Low"                                                                                                                                                                         ,
  #
  "Anthropic"                        , "Claude Sonnet 4.6"                  , "claude-sonnet-4-6"                                                                                                                                                           ,
  "1M"                               , "128k"                               ,
  "$3"                               , "$15"                                ,
  "Aug 2025"                         , "Jan 2026"                           , "Moderate"                                                                                                                                                                    ,
  #
  "Anthropic"                        , "Claude Opus 4.7"                    , "claude-opus-4-7"                                                                                                                                                             ,
  "context_window_k"                 , "max_output_k"                       ,
  "cost_input_mtok"                  , "cost_output_mtok"                   ,
  "knowledge_cutoff"                 , "training_cutoff"                    , "High"                                                                                                                                                                        ,
  #
  "Anthropic"                        , "Claude Opus 4.8"                    , "claude-opus-4-8"                                                                                                                                                             ,
  "1M"                               , "128k"                               ,
  "$5"                               , "$25"                                ,
  "Jan 2026"                         , "Jan 2026"                           , "High"                                                                                                                                                                        ,
  # ---- OpenAI ------------------------------------------------------------------
  # TODO: OpenAI distinguishes between input and cached input cost. Wwhat does this mean? I used regular input for now.
  # Also only provides "knowledge cutoff"
  "OpenAI"                           , "GPT-5.4 Mini"                       , "gpt-5.4-mini"                                                                                                                                                                ,
  "400"                              , "128"                                ,
  "0.75"                             , "4.50"                               ,
  "Aug 2025"                         , "training_cutoff"                    , "Our strongest mini model yet for coding, computer use, and subagents"                                                                                                        ,
  #
  "OpenAI"                           , "GPT-5.4"                            , "gpt-5.4"                                                                                                                                                                     ,
  "1050"                             , "128"                                ,
  "2.5"                              , "15.0"                               ,
  "Aug 2025"                         , "training_cutoff"                    , "A more affordable model for coding and professional work."                                                                                                                   ,
  #
  "OpenAI"                           , "GPT-5.5"                            , "gpt-5.5"                                                                                                                                                                     ,
  "1050"                             , "128"                                ,
  "cost_input_mtok"                  , "30.0"                               ,
  "Dec 2025"                         , "training_cutoff"                    , "A new class of intelligence for coding and professional work."                                                                                                               ,
  # ---- Google Gemini -----------------------------------------------------------
  "Google Gemini"                    , "Gemini 2.5 Pro"                     , "gemini-2.5-pro"                                                                                                                                                              ,
  "1050"                             , "65"                                 ,
  "1.25 (<= 200k tokens), then 2.5"  , "10.00 (<= 200k tokens), then 15.00" ,
  "Jan 2025"                         , "training_cutoff"                    , "Our state-of-the-art multipurpose model, which excels at coding and complex reasoning tasks."                                                                                ,
  #
  "Google Gemini"                    , "Gemini 3.1 Pro Preview"             , "gemini-3.1-pro-preview"                                                                                                                                                      ,
  "context_window_k"                 , "max_output_k"                       ,
  "2.00 (<= 200k tokens), then 4.00" , "12.00 (<= 200k tokens), then 18.00" ,
  "knowledge_cutoff"                 , "training_cutoff"                    , "The latest performance, intelligence, and usability improvements to the best model family in the world for multimodal understanding, agentic capabilities, and vibe-coding." ,
  #
  "Google Gemini"                    , "Gemini 3.5 Flash"                   , "gemini-3.5-flash"                                                                                                                                                            ,
  "1000"                             , "65"                                 ,
  "1.50"                             , "9.00"                               ,
  "Jan 2025"                         , "training_cutoff"                    , "Our most intelligent model built for speed, combining frontier intelligence with superior search and grounding."
)
# Named vector for use in experimental_setup (label -> API id)
models <- setNames(models_tbl$model_id, models_tbl$model_label)


models_tbl |> flextable() |> autofit()
```

## API Calls

Provider APIs are called using the existing API call functionality developed for eData (`STOPeData::extract_pdf_with_llm()`). Each call returns a list object containing extracted data and call metadata (cost, tokens, etc.), which is then saved as as an `.rds` file. 

```{r}
#| label: API-calls-setup
#| code-fold: true

# Recursively walk the list, collecting path-as-vector as we go.
# Each branch terminates at: (1) a data frame, (2) an atomic scalar.
flatten_to_rows <- function(x, path = character()) {
  if (is.data.frame(x)) {
    # --- TIBBLE BRANCH ---
    # Capture column types before pivoting (pivot coerces everything to one type)
    col_types <- sapply(x, \(col) paste(class(col), collapse = "/"))

    x |>
      mutate(.row_idx = row_number()) |>
      pivot_longer(
        cols = -.row_idx,
        names_to = "name",
        values_to = "value",
        values_transform = as.character # coerce all cells → chr for uniform value col
      ) |>
      mutate(
        level_path = list(path), # store the full path as a list-col
        row_idx = .row_idx,
        type = col_types[name] # look up original type by col name
      ) |>
      select(level_path, name, row_idx, value, type)
  } else if (is.list(x)) {
    # --- LIST BRANCH: recurse into each named element ---
    imap(x, \(item, nm) flatten_to_rows(item, c(path, nm))) |>
      list_rbind()
  } else {
    # --- SCALAR BRANCH ---
    # The last segment of path IS the name; everything before is the path.
    tibble(
      level_path = list(head(path, -1)),
      name = tail(path, 1),
      row_idx = NA_integer_,
      value = as.character(x),
      type = paste(class(x), collapse = "/")
    )
  }
}

# Map provider display names → the ellmer chat function and API key env var.
# model_provider is the string extract_pdf_with_llm() checks internally.
provider_config <- tribble(
  ~provider       , ~model_provider , ~env_var            , ~chat_fn             ,
  "Anthropic"     , "Anthropic"     , "ANTHROPIC_API_KEY" , "chat_anthropic"     ,
  "OpenAI"        , "OpenAI"        , "OPENAI_API_KEY"    , "chat_openai"        ,
  "Google Gemini" , "Google"        , "GOOGLE_API_KEY"    , "chat_google_gemini"
)

# Run a single extraction. Looks up the API key from the env var at call time.
run_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn
) {
  extract_pdf_with_llm(
    pdf_path = pdf_path,
    model_provider = model_provider,
    model_name = model_id,
    env_var = env_var,
    chat_fn = chat_fn,
    api_key = Sys.getenv(env_var),
    extraction_prompt = create_extraction_prompt(),
    extraction_schema = create_extraction_schema(),
    max_tokens = as.integer(max_tokens)
  )
}

# Unpack the level_path list-col into level_1 / level_2 character columns.
flatten_call <- function(call_result) {
  flatten_to_rows(call_result) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)
}

# --- Per-call RDS caching ----------------------------------------------------

# Build a deterministic RDS path from the call parameters.
call_rds_path <- function(
  pdf,
  model_id,
  tokens,
  replicate,
  cache_dir = LLM_CACHE_DIR
) {
  # Replace path-unsafe characters in model_id (e.g. "/")
  safe_model <- gsub("/", "-", model_id)
  file.path(
    cache_dir,
    glue("{pdf}_{safe_model}_{tokens}_rep{replicate}.rds")
  )
}

# Run a single extraction, loading from RDS cache if already done.
run_or_load_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn,
  pdf,
  replicate
) {
  rds_path <- call_rds_path(pdf, model_id, max_tokens, replicate)

  if (file.exists(rds_path)) {
    message(glue("Cache hit — loading: {basename(rds_path)}"))
    result <- readRDS(rds_path)
  } else {
    message(glue("Cache miss — running API call: {basename(rds_path)}"))
    result <- run_extraction(
      pdf_path,
      model_id,
      max_tokens,
      model_provider,
      env_var,
      chat_fn
    )
    saveRDS(result, rds_path)
  }
  result
}


# Build an experiment grid and run/load all extractions.
# Each call is cached individually as an RDS keyed on its parameters.
# Returns a tibble with a `call` list-column holding list(result, metadata, success).
#
# pdf_ids:      character vector of PDF ids from the pdfs tibble (e.g. c("a", "b"))
# model_ids:    character vector of model IDs from models_tbl$model_id
# token_counts: integer vector of max_tokens values to test
# n_replicates: how many times to repeat each combination
run_extraction_batch <- function(
  pdf_ids,
  model_ids,
  token_counts,
  n_replicates
) {
  expand.grid(
    pdf = pdf_ids,
    models = model_ids,
    tokens = token_counts,
    replicate = seq_len(n_replicates),
    stringsAsFactors = FALSE
  ) |>
    as_tibble() |>
    left_join(pdfs |> select(id, path), by = c("pdf" = "id")) |>
    left_join(
      models_tbl |>
        select(model_id, provider) |>
        left_join(provider_config, by = "provider"),
      by = c("models" = "model_id")
    ) |>
    mutate(
      call = pmap(
        list(
          pdf_path = path,
          model_id = models,
          max_tokens = tokens,
          model_provider = model_provider,
          env_var = env_var,
          chat_fn = chat_fn,
          pdf = pdf,
          replicate = replicate
        ),
        run_or_load_extraction
      )
    )
}

# Rebuild the flat parquet from all per-call RDS files present in the cache.
# Called every time — cheap, since it just reads already-saved RDS files.
rebuild_flat_parquet <- function(
  calls_results,
  parquet_path = here("data-raw/llm_cache/llm_eval_output_flat.parquet")
) {
  calls_flat <- calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      id = as.character(row_number()),
      flat = map(call, flatten_call)
    ) |>
    select(id, pdf, models, tokens, replicate, flat) |>
    unnest(flat)

  write_parquet(calls_flat, parquet_path)
  calls_flat
}
```

For each manuscript and model multiple calls were made, with token count was set to 3,000, 6,000 and 9,000 to determine if call success, effectiveness, and quality varied with this. Each combination of call, model, token count, and pdf was made 5 times, to provide a measure of inter- and intra-call variation.

```{r}
#| label: tbl-experimental-design
#| tbl-cap: "Experimental design: combinations of PDFs, models, and token limits."

tibble(
  Parameter = c(
    "PDF IDs",
    "Model IDs",
    "Token Limits",
    "Replicates per Combination"
  ),
  Values = c(
    paste(c("a", "b", "c", "d"), collapse = ", "),
    paste(c("claude-haiku-4-5", "claude-sonnet-4-6"), collapse = ", "),
    paste(c(3000, 6000, 9000), collapse = ", "),
    "5"
  )
) |>
  flextable() |>
  autofit() |>
  set_header_labels(Parameter = "Parameter", Values = "Values")
```

```{r}
#| label: api-calls

calls_results <- run_extraction_batch(
  pdf_ids = c("a"),
  model_ids = c("claude-haiku-4-5", "claude-sonnet-4-6"),
  token_counts = c(3000, 6000),
  n_replicates = 3
) |>
  mutate(id = row_number(), .before = everything())

# nb: the number of IDs in calls_results and calls_flats may not be the same, because the flat file also includes all previous calls in the folder
calls_flat <- rebuild_flat_parquet(calls_results)

# Named list of individual call objects, keyed by their parameter combination.
# Scales automatically with whatever is in calls_results.
calls <- setNames(
  calls_results$call,
  with(
    calls_results,
    paste(pdf, models, tokens, paste0("rep", replicate), sep = "_")
  )
)

# --- Empty-table detection ---------------------------------------------------
# "Successful" API calls (success == TRUE) may still return all-NA tables if
# the token limit was hit mid-response. Identify these cases here.
# A table is "empty" if every value in it is NA.

call_completeness <- calls_flat |>
  group_by(id, table = level_2) |>
  summarise(
    n_non_na = sum(!is.na(value)),
    is_empty = n_non_na == 0,
    .groups = "drop"
  ) |>
  group_by(id) |>
  summarise(
    tables_with_data = sum(!is_empty),
    tables_empty = sum(is_empty),
    .groups = "drop"
  )

call_status <- calls_results |>
  mutate(
    id = as.character(row_number()),
    api_success = map_lgl(call, \(c) isTRUE(c$success))
  ) |>
  left_join(call_completeness, by = "id") |>
  replace_na(list(tables_with_data = 0L, tables_empty = 0L)) |>
  mutate(
    effectively_complete = api_success &
      tables_empty == 0 &
      tables_with_data > 0
  ) |>
  select(
    pdf,
    models,
    tokens,
    replicate,
    api_success,
    tables_with_data,
    tables_empty,
    effectively_complete
  )
```

# Results

## Call Status

[@tbl-batch-report] reports the details and status of each call. Broadly speaking, two types of errors are possible. In case of connection issue, system downtime, or a lack of API credit, the call will return an HTTP error (e.g. HTTP 400), and no data. However, it has also been observed that in some cases, the API response will be flagged "successful" by the external provider, but returned only empty data. This is believed to be the result of the extraction process hitting the token limit before completing its response; for example when an unusually large PDF is send with a too-small token limit. This sharp distinction between performance at different token limits 

::: {.callout-note collapse="true"}
## Per-Replicate Call Status Raw Data

```{r}
#| label: tbl-batch-report
#| tbl-cap: "Per-replicate API call status and error messages."

# --- Batch run status report -------------------------------------------------
# Inspect the call list-column directly — before any flattening — to report
# on successes and errors with token counts, broken down by pdf, model, and
# token limit.

batch_run_report <- calls_results |>
  mutate(
    success = map_lgl(call, \(c) isTRUE(c$success)),
    error_msg = map_chr(call, \(c) c$error %||% NA_character_),
    # metadata is a named list of summed token counts for successful calls.
    # Try both naming conventions used across ellmer versions.
    input_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["input_tokens"]] %||% c$metadata[["input"]] %||% NA_real_
        ),
        error = \(e) NA_integer_
      )
    }),
    output_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["output_tokens"]] %||%
            c$metadata[["output"]] %||%
            NA_real_
        ),
        error = \(e) NA_integer_
      )
    })
  )

# Per-replicate detail
batch_run_report |>
  select(
    pdf,
    models,
    tokens,
    replicate,
    success,
    error_msg
  ) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    replicate = "Rep",
    success = "Success",
    error_msg = "Error"
  ) |>
  color(i = ~ !success, j = ~ success + error_msg, color = "firebrick")
```

:::

```{r}
#| label: tbl-batch-report-summary
#| tbl-cap: "Summary of API call success rates and mean token usage by PDF, model, and token limit."

# Grouped summary: counts and mean token usage by pdf × model × token limit
batch_run_report |>
  group_by(pdf, models, tokens) |>
  summarise(
    n_calls = n(),
    n_success = sum(success),
    n_failed = sum(!success),
    error_types = paste(sort(unique(na.omit(error_msg))), collapse = "; "),
    mean_input_tok = round(mean(input_tokens, na.rm = TRUE)),
    mean_output_tok = round(mean(output_tokens, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    n_calls = "N",
    n_success = "Success",
    n_failed = "Failed",
    error_types = "Error Types",
    mean_input_tok = "Mean Input Tok",
    mean_output_tok = "Mean Output Tok"
  )
```

::: {.callout-note collapse="true"}
## All Calls Raw Data

```{r}
#| label: tbl-all-calls
#| tbl-cap: "Status of all API calls, including empty-table detection."

call_status |> flextable() |> autofit()
```

:::


```{r}
#| label: tbl-all-calls-summary
#| tbl-cap: "Call completeness summary by PDF, model, and token limit."

call_status |>
  group_by(pdf, models, tokens) |>
  summarise(
    n_calls = n(),
    n_complete = sum(effectively_complete),
    n_empty_tables = sum(api_success & !effectively_complete),
    n_failed = sum(!api_success),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit()
```


## eData Validation

We have functions for this, but they're intended to deal with finished data, not fresh from extraction. Still

```{r}
#| label: edata-validation-setup

# One function per table. Each takes the raw LLM-extracted data frame,
# adds dummy values for columns the LLM does not generate, and returns an
# interrogated pointblank agent. Call get_agent_report(agent) to inspect
# any individual table in detail.
#
# Dummied columns are noted inline. Steps that check generated fields
# (SAMPLE_ID, PROTOCOL_ID) will fail — this is expected.
#
# Tables skipped:
#   samples  — LLM schema (sampling_dates, sampling_site_code, …) does not
#              map to the eData Samples table.
#   comments — No pb_validate_*() function exists for this table.

validate_llm_campaign <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(ENTERED_BY = "Dummy", ENTERED_DATE = today()) |>
    pb_validate_campaign(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_references <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      REFERENCE_TYPE = "Journal Article",
      REFERENCE_ID = generate_reference_id(YEAR, AUTHOR, TITLE),
      ACCESS_DATE = today(),
      INSTITUTION = "Dummy", # not extracted by LLM
      DATA_SOURCE = "Primary" # not extracted by LLM
    ) |>
    pb_validate_reference(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_sites <- function(tbl) {
  # SITE_GEOGRAPHIC_FEATURE will likely fail: the LLM returns descriptive
  # text rather than a controlled-vocabulary value.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ALTITUDE_VALUE = 0, # not extracted by LLM
      ALTITUDE_UNIT = "km", # not extracted by LLM
      SITE_COORDINATE_SYSTEM = "Not relevant", # not extracted by LLM
      ENTERED_BY = "Dummy",
      ENTERED_DATE = today()
    ) |>
    pb_validate_sites(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_parameters <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ENTERED_BY = "Dummy",
      MEASURED_TYPE = "Concentration" # not extracted by LLM
    ) |>
    pb_validate_parameters(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_compartments <- function(tbl) {
  # No pb_validate_compartments() in eDataDRF; build the agent inline.
  tbl |>
    rename_with(str_to_upper) |>
    create_agent(label = "eData Compartments Validation") |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT,
      set = environ_compartments_vocabulary()
    ) |>
    col_vals_in_set(
      columns = MEASURED_CATEGORY,
      set = unname(measured_categories_vocabulary())
    ) |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT_SUB,
      set = environ_compartments_sub_vocabulary() |> flatten()
    ) |>
    interrogate(progress = FALSE)
}

validate_llm_biota <- function(tbl) {
  # SITE_CODE, PARAMETER_NAME, ENVIRON_COMPARTMENT, ENVIRON_COMPARTMENT_SUB
  # are not extracted by the LLM and are dummied.
  # SAMPLE_ID will fail the regex check — it is a generated field.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      SITE_CODE = "Dummy",
      PARAMETER_NAME = "Dummy",
      ENVIRON_COMPARTMENT = "Biota",
      ENVIRON_COMPARTMENT_SUB = first(
        environ_compartments_sub_vocabulary()$Biota
      )
    ) |>
    pb_validate_biota(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# TODO: Methods is definitely still a problem.
validate_llm_methods <- function(tbl) {
  # PROTOCOL_ID and CAMPAIGN_NAME are not extracted by the LLM and are dummied.
  # PROTOCOL_ID will fail the regex check — it is a generated field.
  # TODO: PROTOCOL_NAME is failing all time time
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      PROTOCOL_ID = "Dummy",
      CAMPAIGN_NAME = "Dummy"
    ) |>
    pb_validate_methods(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# Thin wrapper: dispatches to each per-table function and returns a named
# list of interrogated agents. NULL tables are silently skipped.
validate_llm_extraction <- function(result) {
  fns <- list(
    campaign = validate_llm_campaign,
    references = validate_llm_references,
    sites = validate_llm_sites,
    parameters = validate_llm_parameters,
    compartments = validate_llm_compartments,
    biota = validate_llm_biota,
    methods = validate_llm_methods
  )
  imap(fns, \(fn, nm) {
    tbl <- result[[nm]]
    if (is.null(tbl)) {
      return(NULL)
    }
    if (nrow(tbl) == 0) {
      return(list(.empty = TRUE))
    }
    fn(tbl)
  }) |>
    compact()
}

# Summarise failure counts across all tables in a validation list.
# Returns one row per table with step-level and row-level failure counts.
summarise_validation <- function(validation_list) {
  imap(validation_list, \(agent, table_name) {
    if (isTRUE(agent$.empty)) {
      return(tibble(
        table = table_name,
        n_steps = 0L,
        n_steps_failed = 0L,
        rows_failed = 0L,
        pct_steps_pass = 0
      ))
    }
    agent$validation_set |>
      summarise(
        table = table_name,
        n_steps = n(),
        n_steps_failed = sum(n_failed > 0, na.rm = TRUE),
        rows_failed = sum(n_failed, na.rm = TRUE),
        pct_steps_pass = mean(all_passed, na.rm = TRUE)
      )
  }) |>
    list_rbind()
}

# Run validation across all successful calls and return a long tidy tibble:
# one row per (call × table), with pdf / models / tokens / replicate attached.
#
# For an overall pass rate per call:
#   validate_all_calls(calls_results) |>
#     group_by(pdf, models, tokens, replicate) |>
#     summarise(pct_pass = mean(pct_steps_pass))
validate_all_calls <- function(calls_results) {
  calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      validation_summary = map(call, \(c) {
        validate_llm_extraction(c$result) |>
          summarise_validation()
      })
    ) |>
    select(id, pdf, models, tokens, replicate, validation_summary) |>
    unnest(validation_summary)
}
```


```{r}
#| label: validation-execute

# Keep named agents available for get_agent_report() inspection.
# Index by row-number ID from batch_run_report, e.g. call_validations[["42"]]

# Reuse batch_run_report to identify which calls have a result worth validating.
successful_ids <- batch_run_report$id[batch_run_report$success]
call_validations <- calls[successful_ids] |>
  set_names(successful_ids) |>
  map(\(c) validate_llm_extraction(c$result))

table_summary <- validate_all_calls(calls_results)
```


```{r}
#| label: fig-validation-scores
#| fig-cap: "Proportion of validation steps passed by table and model. Each dot represents one replicate; dots stack at exact values to reflect the discrete nature of step-pass proportions."

table_summary |>
  mutate(tokens = factor(tokens)) |>
  ggplot(aes(x = table, y = pct_steps_pass, fill = tokens)) +
  geom_dotplot(
    binaxis = "y",
    stackdir = "center",
    binwidth = 0.04,
    dotsize = 0.5,
    position = position_dodge(width = 0.5)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    name = "Proportion of Validation Passed"
  ) +
  facet_wrap(vars(models))
```

Across all tables, is there a relationship between token count and validation? [@fig-validation-relationship]. Not as present, but there may be with other values.

```{r}
#| label: fig-validation-relationship
#| fig-cap: "Across all tables, is there a relationship between token count and validation?"

table_summary |>
  ggplot(aes(colour = models, y = pct_steps_pass, x = tokens)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_y_continuous(
    limits = c(0, 1),
    name = "Proportion of Validation Passed"
  )
```

```{r}
#| label: canonical-data-setup

pdf_zip_map <- c(
  a = here("data-raw/1997RitterhoffTraceMetalsIn_AllData_20251030_145522.zip"),
  b = here(
    "data-raw/2002HellandTransportAndSedimentation_AllData_20251104_075533.zip"
  ),
  c = here(
    "data-raw/2021WojtasikTemporalIsolationBetween_AllData_20251030_163338.zip"
  ),
  d = here(
    "data-raw/2022KohlerDistributionPatternOf_AllData_20251217_115443.zip"
  )
)

load_canonical <- function(zip_path) {
  tmp <- tempfile()
  dir.create(tmp)
  unzip(zip_path, exdir = tmp)
  csvs <- list.files(tmp, full.names = TRUE, pattern = "\\.csv$")

  patterns <- c(
    campaign = "Campaign",
    references = "Reference",
    sites = "Sites",
    parameters = "Parameters",
    compartments = "Compartments",
    biota = "Biota",
    methods = "Methods"
  )

  imap(patterns, function(pat, nm) {
    f <- csvs[grepl(pat, csvs, fixed = TRUE)]
    if (length(f) == 0) {
      return(NULL)
    }
    read_csv(f[1], show_col_types = FALSE) |> rename_with(tolower)
  }) |>
    compact()
}

# Load canonical data for all four papers (cached in-memory)
canonical_data <- imap(pdf_zip_map, \(path, id) load_canonical(path))
```


```{r}
#| label: canon-comparison-setup
# --- Variable classification --------------------------------------------------
# var_type:
#   "controlled" — compare with case-insensitive exact match → boolean
#   "continuous" — compute numeric delta (LLM − canonical)
#   "exclude"  — excluded from comparison
#
# canonical_name uses lowercase column names (matching rename_with(tolower) above).
# LLM schema names that differ from canonical names are mapped explicitly
# (e.g. country_iso → country, ocean_iho → area).

variable_meta <- tribble(
  ~table         , ~llm_name                     , ~canonical_name               , ~var_type    ,
  # campaign
  "campaign"     , "campaign_start_date"         , "campaign_start_date"         , "continuous" ,
  "campaign"     , "campaign_end_date"           , "campaign_end_date"           , "continuous" ,
  "campaign"     , "campaign_name"               , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_name_short"         , NA_character_                 , "exclude"    ,
  "campaign"     , "organisation"                , "organisation"                , "exclude"    ,
  "campaign"     , "campaign_comment"            , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_rationale"          , NA_character_                 , "exclude"    ,
  # references
  "references"   , "reference_type"              , "reference_type"              , "controlled" ,
  "references"   , "year"                        , "year"                        , "continuous" ,
  "references"   , "volume"                      , "volume"                      , "continuous" ,
  "references"   , "issue"                       , "issue"                       , "continuous" ,
  "references"   , "author"                      , NA_character_                 , "exclude"    ,
  "references"   , "title"                       , NA_character_                 , "exclude"    ,
  "references"   , "doi"                         , NA_character_                 , "exclude"    ,
  "references"   , "periodical_journal"          , NA_character_                 , "exclude"    ,
  "references"   , "publisher"                   , NA_character_                 , "exclude"    ,
  # sites
  "sites"        , "latitude"                    , "latitude"                    , "continuous" ,
  "sites"        , "longitude"                   , "longitude"                   , "continuous" ,
  "sites"        , "country_iso"                 , "country"                     , "controlled" ,
  "sites"        , "ocean_iho"                   , "area"                        , "controlled" ,
  "sites"        , "site_geographic_feature"     , "site_geographic_feature"     , "controlled" ,
  "sites"        , "site_geographic_feature_sub" , "site_geographic_feature_sub" , "controlled" ,
  "sites"        , "site_code"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_name"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_comment"                , NA_character_                 , "exclude"    ,
  # parameters
  "parameters"   , "parameter_type"              , "parameter_type"              , "controlled" ,
  "parameters"   , "cas_rn"                      , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_name"              , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_comment"           , NA_character_                 , "exclude"    ,
  # compartments
  "compartments" , "environ_compartment"         , "environ_compartment"         , "controlled" ,
  "compartments" , "environ_compartment_sub"     , "environ_compartment_sub"     , "controlled" ,
  "compartments" , "measured_category"           , "measured_category"           , "controlled" ,
  # biota
  "biota"        , "species_group"               , "species_group"               , "controlled" ,
  "biota"        , "sample_tissue"               , "sample_tissue"               , "controlled" ,
  "biota"        , "sample_species_lifestage"    , "sample_species_lifestage"    , "controlled" ,
  "biota"        , "sample_species_gender"       , "sample_species_gender"       , "controlled" ,
  "biota"        , "sample_species"              , NA_character_                 , "exclude"    ,
  "biota"        , "sample_id"                   , NA_character_                 , "exclude"    ,
  # methods
  "methods"      , "protocol_category"           , "protocol_category"           , "controlled" ,
  "methods"      , "protocol_name"               , "protocol_name"               , "controlled" ,
  "methods"      , "protocol_comment"            , NA_character_                 , "exclude"    ,
)

table_keys <- list(
  campaign = NULL,
  references = NULL,
  sites = c("latitude", "longitude"),
  parameters = "parameter_name",
  compartments = c("environ_compartment", "environ_compartment_sub"),
  biota = c("sample_species", "sample_tissue"),
  methods = c("protocol_category", "protocol_name")
)
```

[@tbl-canonical-comparison-variables] we check controlled and numerical variables against the "canonical versions" from the source data. This isn't very sophisticated:
- we just check if strings match
- for tables (sites, parameters, etc.) with more than one row we only check that what's in the call table is also in the canonical table (so if the call over-matches we don't flag that as an issue)

```{r}
#| label: tbl-canonical-comparison-variables
#| tbl-cap: "Caption."
#| echo: false

variable_meta |>
  mutate(
    llm_name = toupper(llm_name),
    table = str_to_sentence(table),
    var_type = case_when(
      var_type %in% c("controlled", "continuous") ~ "Compared",
      var_type == "exclude" ~ "Skipped",
      .default = str_to_sentence(var_type)
    )
  ) |>
  select(table, llm_name, var_type) |>
  flextable() |>
  autofit() |>

  set_header_labels(
    table = "Table",
    llm_name = "Variable Name",
    var_type = "Variable Check"
  )
```

::: {.callout-note collapse="true"}
## Canon Comparison Raw Data

```{r}
#| label: canon-comparison-all

# Scale string comparison across all calls × all tables.
#
# For each (call_id, table) pair:
#   1. Align LLM field names to canonical field names via variable_meta.
#   2. Pool all values across rows (ignoring row identity).
#   3. Per field: any LLM value == any canonical value → match.
# Produces one row per (call × table) with n_match, n_total, prop_match.

canonical_comparison <- calls_flat |>
  distinct(id, pdf, models, tokens, replicate) |>
  crossing(
    table = variable_meta |>
      filter(!is.na(var_type), var_type != "exclude") |>
      distinct(table) |>
      pull(table)
  ) |>
  mutate(
    result = pmap(list(id, pdf, table), \(call_id, pdf_id, tbl) {
      canonical_tbl <- canonical_data[[pdf_id]][[tbl]]
      if (is.null(canonical_tbl)) {
        return(NULL)
      }

      meta <- variable_meta |>
        filter(
          .data$table == tbl,
          !is.na(.data$var_type),
          .data$var_type != "exclude"
        )

      llm_vals <- calls_flat |>
        filter(id == call_id, level_2 == tbl) |>
        select(name, value) |>
        inner_join(
          meta |> select(llm_name, canonical_name),
          by = c(name = "llm_name")
        ) |>
        select(canonical_name, value_call = value)

      canon_vals <- canonical_tbl |>
        pivot_longer(cols = everything(), values_transform = as.character) |>
        inner_join(
          meta |> select(canonical_name),
          by = c(name = "canonical_name")
        ) |>
        select(canonical_name = name, value_canon = value)

      if (nrow(llm_vals) == 0 || nrow(canon_vals) == 0) {
        return(NULL)
      }

      # For each field: does any LLM value match any canonical value?
      llm_vals |>
        left_join(
          canon_vals,
          by = "canonical_name",
          relationship = "many-to-many"
        ) |>
        group_by(canonical_name) |>
        summarise(
          match = any(value_call == value_canon, na.rm = TRUE),
          .groups = "drop"
        ) |>
        summarise(
          n_match = sum(match, na.rm = TRUE),
          n_total = n(),
          prop_match = n_match / n_total
        )
    })
  ) |>
  filter(!map_lgl(result, is.null)) |>
  unnest(result)

canonical_comparison |>
  flextable() |>
  autofit() |>
  colformat_double(j = "prop_match", digits = 2) |>
  set_header_labels(
    id = "Call ID",
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    replicate = "Rep",
    table = "Table",
    n_match = "N Matched",
    n_total = "N Total",
    prop_match = "Prop. Matched"
  )
```

:::

Our smaller token calls (3000) end up all excluded here, because with all NA data there's nothing to compare.

```{r}
#| label: fig-canon-comparison-summary
#| fig-cap:
#|   - "Mean proportion of fields matching canonical data, by model and token limit."
#|   - "Proportion of fields matching canonical data, by model and table."
# what's the best way to summarise this?
canonical_comparison_summarised <- canonical_comparison |>
  group_by(models, tokens) |>
  reframe(mean_prop_match = mean(prop_match), n_replicates = max(replicate))

# our questions are:
# which models are most accurate?
# which tables cause the most problems?
# does token count affect anything?

# for a standard "look how good my LLM is" comparison, we'd just do a bar chart...

canonical_comparison_summarised |>
  ggplot(aes(x = models, y = mean_prop_match, fill = tokens)) +
  geom_boxplot() +
  scale_y_continuous(limits = c(0, 1))
```
```{r}
#| label: fig-canon-comparison-summary-2
# realistically, we should at the very least do summary stats
canonical_comparison |>
  ggplot(aes(x = models, y = prop_match, colour = table)) +
  geom_boxplot()

# again, we have the issue that points overplot. better with boxplots (sort of)
```

```{r}
#| label: compare-canonical-comments

# Screening comments are a special case: there is no canonical ground truth
# to compare against, so we assess inter-rater reliability between
# models/replicates using Cohen's / Fleiss' kappa.
#
# Weighted kappa is used throughout: it penalises larger disagreements more
# (e.g. 1 vs 5) than small ones (1 vs 2), which is appropriate for 1–5
# ordinal scores.

score_vars <- c(
  "paper_relevance",
  "paper_reliability",
  "paper_data_source",
  "paper_data_available",
  "extraction_assessement"
)

# Extract integer scores from the "Score: N: text" format
scores_long <- calls_flat |>
  filter(level_2 == "comments", name %in% score_vars) |>
  select(id, pdf, models, tokens, replicate, name, value) |>
  mutate(score = as.integer(str_extract(value, "(?<=Score:\\s)\\d"))) |>
  select(-value)


# --- Within-model kappa (consistency across replicates) ----------------------
# For each (model, score_variable): matrix rows = pdfs, cols = replicates.
# Multiple raters → Fleiss' kappa.

within_model_kappa <- scores_long |>
  group_by(models, name) |>
  group_map(function(df, key) {
    mat <- df |>
      select(pdf, replicate, score) |>
      pivot_wider(names_from = replicate, values_from = score) |>
      select(-pdf) |>
      as.matrix()

    # Drop all-NA columns (failed/empty replicates)
    mat <- mat[, colSums(!is.na(mat)) > 0, drop = FALSE]
    if (ncol(mat) < 2 || nrow(mat) < 2) {
      return(tibble())
    }

    res <- tryCatch(kappam.fleiss(mat), error = function(e) NULL)
    if (is.null(res)) {
      return(tibble())
    }

    tibble(
      models = key$models,
      score_var = key$name,
      n_subjects = nrow(mat),
      n_raters = ncol(mat),
      kappa = round(res$value, 3),
      z = round(res$statistic, 2),
      p_value = round(res$p.value, 4),
      comparison = "within_model (replicates)"
    )
  }) |>
  list_rbind()


# --- Cross-model kappa (agreement between models) ----------------------------
# Average scores across replicates per (pdf, model) first, then compare models.
# 2 models → weighted kappa2; >2 models → Fleiss kappa.

cross_model_kappa <- scores_long |>
  group_by(pdf, models, name) |>
  summarise(mean_score = round(mean(score, na.rm = TRUE)), .groups = "drop") |>
  group_by(name) |>
  group_map(function(df, key) {
    mat <- df |>
      select(pdf, models, mean_score) |>
      pivot_wider(names_from = models, values_from = mean_score) |>
      select(-pdf) |>
      as.matrix()

    mat <- mat[, colSums(!is.na(mat)) > 0, drop = FALSE]
    if (ncol(mat) < 2 || nrow(mat) < 2) {
      return(tibble())
    }

    res <- tryCatch(
      if (ncol(mat) == 2) {
        kappa2(mat, weight = "equal")
      } else {
        kappam.fleiss(mat)
      },
      error = function(e) NULL
    )
    if (is.null(res)) {
      return(tibble())
    }

    tibble(
      models = "all",
      score_var = key$name,
      n_subjects = nrow(mat),
      n_raters = ncol(mat),
      kappa = round(res$value, 3),
      z = round(res$statistic, 2),
      p_value = round(res$p.value, 4),
      comparison = "cross_model"
    )
  }) |>
  list_rbind()
```


```{r}
#| label: tbl-combined-results
#| tbl-cap: "Within-model and cross-model kappa statistics for screening score inter-rater agreement."

# Combined results table
bind_rows(within_model_kappa, cross_model_kappa) |>
  arrange(comparison, score_var, models) |>
  flextable() |>
  autofit()
```


