---
title: "LLM Evaluation"
subtitle: "Comparing the quality, speed, and cost of LLM extraction"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(STOPeData)
library(ellmer)
library(tidyverse)
library(flextable)
library(arrow)
library(sf)
library(eDataDRF)
library(pointblank)

devtools::load_all()
```

The `ellmer` package includes functions that connect to a wide variety of local and remote APIs from a variety of providers. Put simply, we have a great many options. Which are better, and which are worse? We might assume that more expensive models are more effective at data extraction, and that larger and more complex files are more difficult - and require more tokens - than simpler models when extracting data. But can we prove any of these assumptions?

The question of how we evaluate different LLMs is an important one. I'm not quite sure how to do it. There are some obvious measures:

- Time
- Cost
- Does the model fail entirely?
- Does the model hit a token limit, or otherwise fail totally?
- Is the data returned obviously wrong/incomplete?

All of these are easy to measure, but the question of the overall quality and accuracy of extracted data are more complicated. We can run our `eDataDRF` validation functions against whatever we get out, but they won't be a perfect fit as they're designed to work with data after human extraction. Still, it may be the easiest approach.

We definitely want to evaluate our "recommended" models from Anthropic, OpenAI and Google. I don't anticipate any major differences. I'd also like to try some of the most popular models from LM Studio, as it supports PDF data extraction out the box, although I haven't been able to get it working yet. 

I then suggest some sort of the following experimental design:

- 3 providers x 2 models, + 1 LM Studio (n = 7)
- 3 different papers of varying lengths (n = 3)
- 3 different token limits, small, medium, and large (n = 3)
- 3 replicates (n = 3) # increase once we've tested it

This will give us 7 * 27 = 189 cases. Which is quite a lot, but we can probably batch them and let them get on with themselves. 

Use multiple models to do a consensus thing and maybe compare against people doing the data extraction manually

Time to pick some different papers!

```{r}
#| label: pdfs

# TODO: Update with details of new papers

pdfs <- tibble(
  id = c("a", "b", "c", "d"),
  path = c(
    "data-raw/1997RitterhoffTraceMetalsIn_MS.pdf",
    "data-raw/2002HellandTransportAndSedimentation_MS.pdf",
    "data-raw/2021WojtasikTemporalIsolationBetween_MS.pdf",
    "data-raw/2022KohlerDistributionPatternOf_MS.pdf"
  ),
  comment = c(
    "Filler",
    "Filler",
    "Filler",
    "Filler"
  ),
  # Fill in after reviewing each PDF
  num_images = c(0, 0, 0, 0),
  num_tables = c(0, 0, 0, 0)
) |>
  mutate(
    info = map(path, pdftools::pdf_info),
    file_size_kb = map_dbl(path, \(p) file.size(p) / 1024),
    num_pages = map_int(info, \(x) x$pages)
  ) |>
  select(-info)

# TODO: Add more paper details, default rounding
pdfs |>
  flextable() |>
  autofit()
```

Sources for model comparisons: https://platform.claude.com/docs/en/about-claude/models/overview, https://developers.openai.com/api/docs/models/compare, https://ai.google.dev/gemini-api/docs/models


```{r}
#| label: models

# TODO: Fill out list

models_tbl <- tribble(
  ~provider          , ~model_label        , ~model_id                ,
  ~release_date      , ~context_window_k   , ~max_output_k            ,
  ~cost_input_mtok   , ~cost_output_mtok   ,
  ~knowledge_cutoff  , ~training_cutoff    , ~niche                   ,
  # ---- Anthropic ---------------------------------------------------------------
  "Anthropic"        , "Claude Haiku 4.5"  , "claude-haiku-4-5"       ,
  "release_date"     , "200k"              , "64k"                    ,
  "$1"               , "$5"                ,
  "Feb 2025"         , "Jul 2025"          , "niche"                  ,
  #
  "Anthropic"        , "Claude Sonnet 4.6" , "claude-sonnet-4-6"      ,
  "release_date"     , "1M"                , "128k"                   ,
  "$3"               , "$15"               ,
  "Aug 2025"         , "Jan 2026"          , "niche"                  ,
  #
  "Anthropic"        , "Claude Opus 4.7"   , "claude-opus-4-7"        ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Anthropic"        , "Claude Opus 4.8"   , "claude-opus-4-8"        ,
  "release_date"     , "1M"                , "128k"                   ,
  "$5"               , "$25"               ,
  "Jan 2026"         , "Jan 2026"          , "niche"                  ,
  # ---- OpenAI ------------------------------------------------------------------
  # TODO: OpenAI distinguishes between input and cached input cost. Wwhat does this mean? I used regular input for now.
  # Also only provides "knowledge cutoff"
  "OpenAI"           , "GPT-5.4 Mini"      , "gpt-5.4-mini"           ,
  "release_date"     , "400"               , "128"                    ,
  "0.75"             , "4.50"              ,
  "Aug 2025"         , "training_cutoff"   , "niche"                  ,
  #
  "OpenAI"           , "GPT-5.4"           , "gpt-5.4"                ,
  "release_date"     , "1050"              , "128"                    ,
  "2.5"              , "15.0"              ,
  "Aug 2025"         , "training_cutoff"   , "niche"                  ,
  #
  "OpenAI"           , "GPT-5.5"           , "gpt-5.5"                ,
  "release_date"     , "1050"              , "128"                    ,
  "cost_input_mtok"  , "30.0"              ,
  "Dec 2025"         , "training_cutoff"   , "niche"                  ,
  # ---- Google Gemini -----------------------------------------------------------
  "Google Gemini"    , "Gemini 2.5 Pro"    , "gemini-2.5-pro"         ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Google Gemini"    , "Gemini 3.1 Pro"    , "gemini-3.1-pro-preview" ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Google Gemini"    , "Gemini 3.5 Flash"  , "gemini-3.5-flash"       ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"
)
# Named vector for use in experimental_setup (label -> API id)
models <- setNames(models_tbl$model_id, models_tbl$model_label)


models_tbl |> flextable() |> autofit()
```

So, what experiments do we want to run?

```{r}
#| label: experimental_design

experimental_setup <-
  expand.grid(
    pdf = c("a", "b", "c"),
    models = models,
    tokens = c("1000", "5000", "10000"),
    replicate = 1:3
  ) |>
  as_tibble()
```


```{r}
#| label: API_calls

# Recursively walk the list, collecting path-as-vector as we go.
# Each branch terminates at: (1) a data frame, (2) an atomic scalar.
flatten_to_rows <- function(x, path = character()) {
  if (is.data.frame(x)) {
    # --- TIBBLE BRANCH ---
    # Capture column types before pivoting (pivot coerces everything to one type)
    col_types <- sapply(x, \(col) paste(class(col), collapse = "/"))

    x |>
      mutate(.row_idx = row_number()) |>
      pivot_longer(
        cols = -.row_idx,
        names_to = "name",
        values_to = "value",
        values_transform = as.character # coerce all cells → chr for uniform value col
      ) |>
      mutate(
        level_path = list(path), # store the full path as a list-col
        row_idx = .row_idx,
        type = col_types[name] # look up original type by col name
      ) |>
      select(level_path, name, row_idx, value, type)
  } else if (is.list(x)) {
    # --- LIST BRANCH: recurse into each named element ---
    imap(x, \(item, nm) flatten_to_rows(item, c(path, nm))) |>
      list_rbind()
  } else {
    # --- SCALAR BRANCH ---
    # The last segment of path IS the name; everything before is the path.
    tibble(
      level_path = list(head(path, -1)),
      name = tail(path, 1),
      row_idx = NA_integer_,
      value = as.character(x),
      type = paste(class(x), collapse = "/")
    )
  }
}

# Map provider display names → the ellmer chat function and API key env var.
# model_provider is the string extract_pdf_with_llm() checks internally.
provider_config <- tribble(
  ~provider       , ~model_provider , ~env_var            , ~chat_fn             ,
  "Anthropic"     , "Anthropic"     , "ANTHROPIC_API_KEY" , "chat_anthropic"     ,
  "OpenAI"        , "OpenAI"        , "OPENAI_API_KEY"    , "chat_openai"        ,
  "Google Gemini" , "Google"        , "GOOGLE_API_KEY"    , "chat_google_gemini"
)

# Run a single extraction. Looks up the API key from the env var at call time.
run_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn
) {
  extract_pdf_with_llm(
    pdf_path = pdf_path,
    model_provider = model_provider,
    model_name = model_id,
    env_var = env_var,
    chat_fn = chat_fn,
    api_key = Sys.getenv(env_var),
    extraction_prompt = create_extraction_prompt(),
    extraction_schema = create_extraction_schema(),
    max_tokens = as.integer(max_tokens)
  )
}

# Unpack the level_path list-col into level_1 / level_2 character columns.
flatten_call <- function(call_result) {
  flatten_to_rows(call_result) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)
}

# Build an experiment grid and run all extractions, returning a tibble with one
# row per run and a `call` list-column holding the raw list(result, metadata, success).
# TODO: Add better error checking for e.g. invalid pdf paths
# pdf_ids:      character vector of PDF ids from the pdfs tibble (e.g. c("a", "b"))
# model_ids:    character vector of model IDs from models_tbl$model_id
# token_counts: integer vector of max_tokens values to test
# n_replicates: how many times to repeat each combination
run_extraction_batch <- function(
  pdf_ids,
  model_ids,
  token_counts,
  n_replicates
) {
  expand.grid(
    pdf = pdf_ids,
    models = model_ids,
    tokens = token_counts,
    replicate = seq_len(n_replicates),
    stringsAsFactors = FALSE
  ) |>
    as_tibble() |>
    left_join(pdfs |> select(id, path), by = c("pdf" = "id")) |>
    left_join(
      models_tbl |>
        select(model_id, provider) |>
        left_join(provider_config, by = "provider"),
      by = c("models" = "model_id")
    ) |>
    mutate(
      call = pmap(
        list(
          pdf_path = path,
          model_id = models,
          max_tokens = tokens,
          model_provider = model_provider,
          env_var = env_var,
          chat_fn = chat_fn
        ),
        run_extraction
      )
    )
}

# TODO: Cache per call
if (
  !file.exists("data-raw/llm_cache/llm_eval_calls.rds") ||
    !file.exists("data-raw/llm_cache/llm_eval_output_flat.parquet")
) {
  calls_results <- run_extraction_batch(
    pdf_ids = "a",
    model_ids = c("claude-haiku-4-5", "claude-sonnet-4-6"),
    token_counts = 6000,
    n_replicates = 5
  )

  # Save raw results list for validation and per-call inspection
  saveRDS(calls_results, "data-raw/llm_cache/llm_eval_calls.rds")

  # Flatten successful calls; keep an id column for downstream pairwise code
  calls_flat <- calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      id = as.character(row_number()),
      flat = map(call, flatten_call)
    ) |>
    select(id, pdf, models, tokens, replicate, flat) |>
    unnest(flat)

  write_parquet(calls_flat, "data-raw/llm_cache/llm_eval_output_flat.parquet")
} else {
  message("LLM API calls cached, loading from /data-raw/llm_cache/")
  calls_results <- readRDS("data-raw/llm_cache/llm_eval_calls.rds")
  calls_flat <- read_parquet("data-raw/llm_cache/llm_eval_output_flat.parquet")
}

# Expose individual call objects for downstream compatibility
call1 <- calls_results$call[[1]]
call2 <- calls_results$call[[2]]
```


```{r}
#| label: call-comparison
calls_flat

# KET is right, pairwise might get messy quickly

# campaign and reference are easy.

# almost certainly there are better options than identical()
compare_campaign <- calls_flat |>
  filter(level_2 == "campaign") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# no issues with references
compare_references <- calls_flat |>
  filter(level_2 == "references") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# sites...
# issue: different calls return variables in different orders!
# we can't automatically match sites to each other unless we assume that the lat/long are always equal (or, if we have a canonical point of comparison, I guess)
compare_sites <- calls_flat |>
  filter(level_2 == "sites") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# we can presumably compare distance with sf

# for example
compare_sites |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric))

# how do we check if each value has an equivalent?
# I guess we just group and ignore idx
# we may have to call st_distance for g
distances <- calls_flat |>
  filter(level_2 == "sites") |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric)) |>
  group_by(id, row_idx) |>
  pivot_wider(names_from = name, values_from = value) |>
  st_as_sf(coords = c("latitude", "longitude"), na.fail = FALSE)


# pair sites and check (great circle) distance?

x <- distances[1:3, 6]
y <- distances[4:6, 6]

# I guess we make a distance matrix, which should have one value of approximately 0 and be... orthogonal?
# I believe these distances are in km, would be good to check
st_distance(x, y)

# parameters should be easier, they have more-or-less canonical names
compare_parameters <- calls_flat |>
  filter(level_2 == "parameters") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# compartments - easy
compare_compartments <- calls_flat |>
  filter(level_2 == "compartments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# biota
# common or linnean name is acceptable as long as they agree
compare_biota <- calls_flat |>
  filter(level_2 == "biota") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# methods
# again, pairing may be an issue here
compare_methods <- calls_flat |>
  filter(level_2 == "methods") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# samples
# might not do the evaluation on this one, since we don't strictly speaking use it
compare_samples <- calls_flat |>
  filter(level_2 == "samples") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# comments are pleasantly easy
compare_comments <- calls_flat |>
  filter(level_2 == "comments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call") |>
  mutate(
    score_1 = str_extract(call1, "(?!Score\\: )[1-5]"),
    score_2 = str_extract(call2, "(?!Score\\: )[1-5]"),
    .before = call1
  )
```

Now that I come to think of it we do already have canonical versions (even if they were partly LLM-assisted, they've been manually checked.) They're not quite in the perfect format, but we can work with them.

Where does that leave us:

1. Run the extraction (n = 1) for a paper
2. Adapt validation functions to the extracted data (may be easier said than done)
3. Do a second run of one of the papers. Do a pairwise comparison (as above) for consistency. How do we scale to n > 2?
4. Scale...

## eData Validation

We have functions for this, but they're intended to deal with finished data, not fresh from extraction. Still


```{r}
#| label: edata-validation

# One function per table. Each takes the raw LLM-extracted data frame,
# adds dummy values for columns the LLM does not generate, and returns an
# interrogated pointblank agent. Call get_agent_report(agent) to inspect
# any individual table in detail.
#
# Dummied columns are noted inline. Steps that check generated fields
# (SAMPLE_ID, PROTOCOL_ID) will fail — this is expected.
#
# Tables skipped:
#   samples  — LLM schema (sampling_dates, sampling_site_code, …) does not
#              map to the eData Samples table.
#   comments — No pb_validate_*() function exists for this table.

validate_llm_campaign <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(ENTERED_BY = "Dummy", ENTERED_DATE = today()) |>
    pb_validate_campaign(agent = TRUE) |>
    interrogate()
}

validate_llm_references <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      REFERENCE_TYPE = "Journal Article",
      REFERENCE_ID = generate_reference_id(YEAR, AUTHOR, TITLE),
      ACCESS_DATE = today(),
      INSTITUTION = "Dummy", # not extracted by LLM
      DATA_SOURCE = "Primary" # not extracted by LLM
    ) |>
    pb_validate_reference(agent = TRUE) |>
    interrogate()
}

validate_llm_sites <- function(tbl) {
  # SITE_GEOGRAPHIC_FEATURE will likely fail: the LLM returns descriptive
  # text rather than a controlled-vocabulary value.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ALTITUDE_VALUE = 0, # not extracted by LLM
      ALTITUDE_UNIT = "km", # not extracted by LLM
      SITE_COORDINATE_SYSTEM = "Not relevant", # not extracted by LLM
      ENTERED_BY = "Dummy",
      ENTERED_DATE = today()
    ) |>
    pb_validate_sites(agent = TRUE) |>
    interrogate()
}

validate_llm_parameters <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ENTERED_BY = "Dummy",
      MEASURED_TYPE = "Concentration" # not extracted by LLM
    ) |>
    pb_validate_parameters(agent = TRUE) |>
    interrogate()
}

validate_llm_compartments <- function(tbl) {
  # No pb_validate_compartments() in eDataDRF; build the agent inline.
  tbl |>
    rename_with(str_to_upper) |>
    create_agent(label = "eData Compartments Validation") |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT,
      set = environ_compartments_vocabulary()
    ) |>
    col_vals_in_set(
      columns = MEASURED_CATEGORY,
      set = unname(measured_categories_vocabulary())
    ) |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT_SUB,
      set = environ_compartments_sub_vocabulary() |> flatten()
    ) |>
    interrogate()
}

validate_llm_biota <- function(tbl) {
  # SITE_CODE, PARAMETER_NAME, ENVIRON_COMPARTMENT, ENVIRON_COMPARTMENT_SUB
  # are not extracted by the LLM and are dummied.
  # SAMPLE_ID will fail the regex check — it is a generated field.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      SITE_CODE = "Dummy",
      PARAMETER_NAME = "Dummy",
      ENVIRON_COMPARTMENT = "Biota",
      ENVIRON_COMPARTMENT_SUB = first(
        environ_compartments_sub_vocabulary()$Biota
      )
    ) |>
    pb_validate_biota(agent = TRUE) |>
    interrogate()
}

# TODO: Methods is definitely still a problem.
validate_llm_methods <- function(tbl) {
  # PROTOCOL_ID and CAMPAIGN_NAME are not extracted by the LLM and are dummied.
  # PROTOCOL_ID will fail the regex check — it is a generated field.
  # TODO: PROTOCOL_NAME is failing all time time
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      PROTOCOL_ID = "Dummy",
      CAMPAIGN_NAME = "Dummy"
    ) |>
    pb_validate_methods(agent = TRUE) |>
    interrogate()
}

# Thin wrapper: dispatches to each per-table function and returns a named
# list of interrogated agents. NULL tables are silently skipped.
validate_llm_extraction <- function(result) {
  fns <- list(
    campaign = validate_llm_campaign,
    references = validate_llm_references,
    sites = validate_llm_sites,
    parameters = validate_llm_parameters,
    compartments = validate_llm_compartments,
    biota = validate_llm_biota,
    methods = validate_llm_methods
  )
  imap(fns, \(fn, nm) if (!is.null(result[[nm]])) fn(result[[nm]])) |>
    compact()
}

# Summarise failure counts across all tables in a validation list.
# Returns one row per table with step-level and row-level failure counts.
summarise_validation <- function(validation_list) {
  imap(validation_list, \(agent, table_name) {
    agent$validation_set |>
      summarise(
        table = table_name,
        n_steps = n(),
        n_steps_failed = sum(n_failed > 0, na.rm = TRUE),
        rows_failed = sum(n_failed, na.rm = TRUE),
        pct_steps_pass = mean(all_passed, na.rm = TRUE)
      )
  }) |>
    list_rbind()
}

# Run validation across all successful calls and return a long tidy tibble:
# one row per (call × table), with pdf / models / tokens / replicate attached.
#
# For an overall pass rate per call:
#   validate_all_calls(calls_results) |>
#     group_by(pdf, models, tokens, replicate) |>
#     summarise(pct_pass = mean(pct_steps_pass))
validate_all_calls <- function(calls_results) {
  calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      validation_summary = map(call, \(c) {
        validate_llm_extraction(c$result) |>
          summarise_validation()
      })
    ) |>
    select(pdf, models, tokens, replicate, validation_summary) |>
    unnest(validation_summary)
}

# Keep named agents available for get_agent_report() inspection
call1_validation <- validate_llm_extraction(call1$result)
call2_validation <- validate_llm_extraction(call2$result)

table_summary <- validate_all_calls(calls_results)
```

Ok, we have at least some data to play with. What do they look like?
Boring!


```{r}

table_summary |>
  ggplot(aes(x = table, y = pct_steps_pass, fill = models, group = replicate)) +
  geom_col(position = "dodge")
```