---
title: "LLM Evaluation"
subtitle: "Comparing the quality, speed, and cost of LLM extraction"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
execute:
  error: true
---

```{r}
#| label: setup
library(STOPeData)
library(ellmer)
library(tidyverse, mask.ok = c("filter", "lag"))
library(flextable)
library(arrow)
library(sf)
library(eDataDRF)
library(pointblank)
library(irr)


devtools::load_all() # remove once STOPeData is rebuilt

set_flextable_defaults(
  font.size = 10,
  font.family = 'Aptos',
  digits = 2,
  theme_fun = theme_vanilla,
  padding = 1
)

# Create a location to save LLM API calls, if it doesn't exist.
LLM_CACHE_DIR <- "data-raw/llm_cache/calls"
dir.create(LLM_CACHE_DIR, recursive = TRUE, showWarnings = FALSE)
```

The `ellmer` package includes functions that connect to a wide variety of local and remote APIs from a variety of providers. Put simply, we have a great many options. Which are better, and which are worse? We might assume that more expensive models are more effective at data extraction, and that larger and more complex files are more difficult - and require more tokens - than simpler models when extracting data. But can we prove any of these assumptions?

The question of how we evaluate different LLMs is an important one. I'm not quite sure how to do it. There are some obvious measures:

- Time
- Cost
- Does the model fail entirely?
- Does the model hit a token limit, or otherwise fail totally?
- Is the data returned obviously wrong/incomplete?

All of these are easy to measure, but the question of the overall quality and accuracy of extracted data are more complicated. We can run our `eDataDRF` validation functions against whatever we get out, but they won't be a perfect fit as they're designed to work with data after human extraction. Still, it may be the easiest approach.

We definitely want to evaluate our "recommended" models from Anthropic, OpenAI and Google. I don't anticipate any major differences. I'd also like to try some of the most popular models from LM Studio, as it supports PDF data extraction out the box, although I haven't been able to get it working yet. 

I then suggest some sort of the following experimental design:

- 3 providers x 2 models, + 1 LM Studio (n = 7)
- 3 different papers of varying lengths (n = 3)
- 3 different token limits, small, medium, and large (n = 3)
- 3 replicates (n = 3) # increase once we've tested it

This will give us 7 * 27 = 189 cases. Which is quite a lot, but we can probably batch them and let them get on with themselves. 

Use multiple models to do a consensus thing and maybe compare against people doing the data extraction manually


```{mermaid}
%%| label: workflow-diagram
%%| fig-cap: "Overview of the LLM evaluation pipeline."
flowchart TD
    pdfs["PDFs tibble<br>4 papers"]         --> grid
    models["Models tibble<br>3 providers"]  --> grid
    tok["Token limits<br>3k / 6k / 9k<br>n replicates"] --> grid

    grid["Experiment grid<br>run_extraction_batch"]  --> cache

    cache{"RDS cache hit?"}
    cache -- Yes --> load["Load cached RDS"]
    cache -- No  --> api["LLM API call<br>extract_pdf_with_llm"]

    api -- success --> save["Save RDS"]
    api -- failure --> err["Record error<br>HTTP code + message"]

    save  --> results
    load  --> results
    err   --> results

    results["calls_results<br>call list-column"]

    results --> flat["rebuild_flat_parquet<br>Successful calls only"]
    flat    --> parquet[("calls_flat parquet<br>long / flat format")]

    results --> report["batch_run_report<br>success · error · tokens<br>by pdf · model · limit"]

    parquet --> status["call_status<br>success · empty-table<br>effectiveness flags"]
    parquet --> validate["validate_all_calls<br>pointblank agents<br>per table"]
    parquet --> canon["Canonical comparison<br>vs. human-extracted data<br>controlled + continuous vars"]
    parquet --> kappa["Kappa statistics<br>inter-rater reliability<br>within- and cross-model"]
```


```{r}
#| label: pdfs

pdfs <- tibble(
  id = c("a", "b", "c", "d"),
  path = c(
    "data-raw/1997RitterhoffTraceMetalsIn_MS.pdf",
    "data-raw/2002HellandTransportAndSedimentation_MS.pdf",
    "data-raw/2021WojtasikTemporalIsolationBetween_MS.pdf",
    "data-raw/2022KohlerDistributionPatternOf_MS.pdf"
  ),
  title = c(
    "Trace metals in field samples of zooplankton from the Fram Strait and the Greenland Sea",
    "Temporal isolation between two strongly differentiated stocks of the Greenland halibut (<i>Reinhardtius hippoglossoides</i> Walbaum, 1792) from the Western Barents Sea",
    "Transport and sedimentation of Cu in a microtidal estuary, SE Norway",
    "Distribution pattern of mercury in northern Barents Sea and Eurasian Basin surface sediment"
  ),
  year = c(1997L, 2002L, 2021L, 2022L),
  comment = c(
    "Well-OCR'd paper with summary statistics of trace metal concentrations reported across wide range zooplankton taxa. Sampling sites reported in low-quality map. Complex, uneven data structure due to random nature of zooplankton sampling. Methodology reported in moderate detail.",
    "Native PDF.  Sites reported in map. Cu concentrations (and flux) reported from a variety of sediment traps at different timepoints. Methodology reported in moderate detail.",
    "Native PDF. Study primarily concerns evolution but includes heavy metal analysis. Sites reported as linear trawls with start/end coordinates. Raw Cu concentrations given for each individual, reported as table. Methodology reported in high detail.",
    "Native PDF. Sampling sites and Cu concentrations only reported as summaries/graphics in paper, but raw data (Cu concentrations, site coordinates, depth) provided on request and prepended to file."
  ),
  # Fill in after reviewing each PDF
  num_images = c(1, 4, 8, 5),
  num_tables = c(7, 2, 8, 2)
) |>
  mutate(
    info = map(path, pdftools::pdf_info),
    file_size_kb = map_dbl(path, \(p) file.size(p) / 1024),
    num_pages = map_int(info, \(x) x$pages)
  ) |>
  select(-info)

# TODO: Add more paper details
pdfs |>
  select(-path) |>
  flextable() |>
  autofit() |>
  colformat_double(j = c("file_size_kb")) |>
  colformat_int(j = c("year"), big.mark = "") |>
  set_header_labels(
    id = "ID",
    title = "Title",
    year = "Year",
    comment = "Comment",
    num_images = "n Images",
    num_tables = "n Tables",
    file_size_kb = "File Size (kb)",
    num_pages = "n Pages"
  )
```

Sources for model comparisons: https://platform.claude.com/docs/en/about-claude/models/overview, https://developers.openai.com/api/docs/models/compare, https://ai.google.dev/gemini-api/docs/models


```{r}
#| label: models

# TODO: Fill out list

models_tbl <- tribble(
  ~provider          , ~model_label        , ~model_id                ,
  ~release_date      , ~context_window_k   , ~max_output_k            ,
  ~cost_input_mtok   , ~cost_output_mtok   ,
  ~knowledge_cutoff  , ~training_cutoff    , ~complexity              ,
  # ---- Anthropic ---------------------------------------------------------------
  "Anthropic"        , "Claude Haiku 4.5"  , "claude-haiku-4-5"       ,
  "release_date"     , "200k"              , "64k"                    ,
  "$1"               , "$5"                ,
  "Feb 2025"         , "Jul 2025"          , "Low"                    ,
  #
  "Anthropic"        , "Claude Sonnet 4.6" , "claude-sonnet-4-6"      ,
  "release_date"     , "1M"                , "128k"                   ,
  "$3"               , "$15"               ,
  "Aug 2025"         , "Jan 2026"          , "Moderate"               ,
  #
  "Anthropic"        , "Claude Opus 4.7"   , "claude-opus-4-7"        ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "High"                   ,
  #
  "Anthropic"        , "Claude Opus 4.8"   , "claude-opus-4-8"        ,
  "release_date"     , "1M"                , "128k"                   ,
  "$5"               , "$25"               ,
  "Jan 2026"         , "Jan 2026"          , "High"                   ,
  # ---- OpenAI ------------------------------------------------------------------
  # TODO: OpenAI distinguishes between input and cached input cost. Wwhat does this mean? I used regular input for now.
  # Also only provides "knowledge cutoff"
  "OpenAI"           , "GPT-5.4 Mini"      , "gpt-5.4-mini"           ,
  "release_date"     , "400"               , "128"                    ,
  "0.75"             , "4.50"              ,
  "Aug 2025"         , "training_cutoff"   , "Low"                    ,
  #
  "OpenAI"           , "GPT-5.4"           , "gpt-5.4"                ,
  "release_date"     , "1050"              , "128"                    ,
  "2.5"              , "15.0"              ,
  "Aug 2025"         , "training_cutoff"   , "Moderate"               ,
  #
  "OpenAI"           , "GPT-5.5"           , "gpt-5.5"                ,
  "release_date"     , "1050"              , "128"                    ,
  "cost_input_mtok"  , "30.0"              ,
  "Dec 2025"         , "training_cutoff"   , "Moderate"               ,
  # ---- Google Gemini -----------------------------------------------------------
  "Google Gemini"    , "Gemini 2.5 Pro"    , "gemini-2.5-pro"         ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "High"                   ,
  #
  "Google Gemini"    , "Gemini 3.1 Pro"    , "gemini-3.1-pro-preview" ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "High"                   ,
  #
  "Google Gemini"    , "Gemini 3.5 Flash"  , "gemini-3.5-flash"       ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "Low"
)
# Named vector for use in experimental_setup (label -> API id)
models <- setNames(models_tbl$model_id, models_tbl$model_label)


models_tbl |> flextable() |> autofit()
```

So, what experiments do we want to run?


```{r}
#| label: API-calls-setup

# Recursively walk the list, collecting path-as-vector as we go.
# Each branch terminates at: (1) a data frame, (2) an atomic scalar.
flatten_to_rows <- function(x, path = character()) {
  if (is.data.frame(x)) {
    # --- TIBBLE BRANCH ---
    # Capture column types before pivoting (pivot coerces everything to one type)
    col_types <- sapply(x, \(col) paste(class(col), collapse = "/"))

    x |>
      mutate(.row_idx = row_number()) |>
      pivot_longer(
        cols = -.row_idx,
        names_to = "name",
        values_to = "value",
        values_transform = as.character # coerce all cells → chr for uniform value col
      ) |>
      mutate(
        level_path = list(path), # store the full path as a list-col
        row_idx = .row_idx,
        type = col_types[name] # look up original type by col name
      ) |>
      select(level_path, name, row_idx, value, type)
  } else if (is.list(x)) {
    # --- LIST BRANCH: recurse into each named element ---
    imap(x, \(item, nm) flatten_to_rows(item, c(path, nm))) |>
      list_rbind()
  } else {
    # --- SCALAR BRANCH ---
    # The last segment of path IS the name; everything before is the path.
    tibble(
      level_path = list(head(path, -1)),
      name = tail(path, 1),
      row_idx = NA_integer_,
      value = as.character(x),
      type = paste(class(x), collapse = "/")
    )
  }
}

# Map provider display names → the ellmer chat function and API key env var.
# model_provider is the string extract_pdf_with_llm() checks internally.
provider_config <- tribble(
  ~provider       , ~model_provider , ~env_var            , ~chat_fn             ,
  "Anthropic"     , "Anthropic"     , "ANTHROPIC_API_KEY" , "chat_anthropic"     ,
  "OpenAI"        , "OpenAI"        , "OPENAI_API_KEY"    , "chat_openai"        ,
  "Google Gemini" , "Google"        , "GOOGLE_API_KEY"    , "chat_google_gemini"
)

# Run a single extraction. Looks up the API key from the env var at call time.
run_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn
) {
  extract_pdf_with_llm(
    pdf_path = pdf_path,
    model_provider = model_provider,
    model_name = model_id,
    env_var = env_var,
    chat_fn = chat_fn,
    api_key = Sys.getenv(env_var),
    extraction_prompt = create_extraction_prompt(),
    extraction_schema = create_extraction_schema(),
    max_tokens = as.integer(max_tokens)
  )
}

# Unpack the level_path list-col into level_1 / level_2 character columns.
flatten_call <- function(call_result) {
  flatten_to_rows(call_result) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)
}

# --- Per-call RDS caching ----------------------------------------------------

# Build a deterministic RDS path from the call parameters.
call_rds_path <- function(
  pdf,
  model_id,
  tokens,
  replicate,
  cache_dir = LLM_CACHE_DIR
) {
  # Replace path-unsafe characters in model_id (e.g. "/")
  safe_model <- gsub("/", "-", model_id)
  file.path(
    cache_dir,
    glue("{pdf}_{safe_model}_{tokens}_rep{replicate}.rds")
  )
}

# Run a single extraction, loading from RDS cache if already done.
run_or_load_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn,
  pdf,
  replicate
) {
  rds_path <- call_rds_path(pdf, model_id, max_tokens, replicate)

  if (file.exists(rds_path)) {
    message(glue("Cache hit — loading: {basename(rds_path)}"))
    result <- readRDS(rds_path)
  } else {
    message(glue("Cache miss — running API call: {basename(rds_path)}"))
    result <- run_extraction(
      pdf_path,
      model_id,
      max_tokens,
      model_provider,
      env_var,
      chat_fn
    )
    saveRDS(result, rds_path)
  }
  result
}


# Build an experiment grid and run/load all extractions.
# Each call is cached individually as an RDS keyed on its parameters.
# Returns a tibble with a `call` list-column holding list(result, metadata, success).
#
# pdf_ids:      character vector of PDF ids from the pdfs tibble (e.g. c("a", "b"))
# model_ids:    character vector of model IDs from models_tbl$model_id
# token_counts: integer vector of max_tokens values to test
# n_replicates: how many times to repeat each combination
run_extraction_batch <- function(
  pdf_ids,
  model_ids,
  token_counts,
  n_replicates
) {
  expand.grid(
    pdf = pdf_ids,
    models = model_ids,
    tokens = token_counts,
    replicate = seq_len(n_replicates),
    stringsAsFactors = FALSE
  ) |>
    as_tibble() |>
    left_join(pdfs |> select(id, path), by = c("pdf" = "id")) |>
    left_join(
      models_tbl |>
        select(model_id, provider) |>
        left_join(provider_config, by = "provider"),
      by = c("models" = "model_id")
    ) |>
    mutate(
      call = pmap(
        list(
          pdf_path = path,
          model_id = models,
          max_tokens = tokens,
          model_provider = model_provider,
          env_var = env_var,
          chat_fn = chat_fn,
          pdf = pdf,
          replicate = replicate
        ),
        run_or_load_extraction
      )
    )
}

# Rebuild the flat parquet from all per-call RDS files present in the cache.
# Called every time — cheap, since it just reads already-saved RDS files.
# TODO: This seems to be doing something strange to our results and flagging calls as successful when they aren't?

# Right, duhh. We only save successful calls to the parquet
rebuild_flat_parquet <- function(
  calls_results,
  parquet_path = "data-raw/llm_cache/llm_eval_output_flat.parquet"
) {
  calls_flat <- calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      id = as.character(row_number()),
      flat = map(call, flatten_call)
    ) |>
    select(id, pdf, models, tokens, replicate, flat) |>
    unnest(flat)

  write_parquet(calls_flat, parquet_path)
  calls_flat
}
```

```{r}
#| label: api-calls

# --- Run / load batch and rebuild parquet ------------------------------------

# TODO: Better error checking. Currently when it fails, you don't find out until you dig into the output

calls_results <- run_extraction_batch(
  pdf_ids = c("a"),
  model_ids = c("claude-haiku-4-5", "claude-sonnet-4-6"),
  token_counts = c(6000),
  n_replicates = 3
) |>
  mutate(id = row_number(), .before = everything())

# nb: the number of IDs in calls_results and calls_flats may not be the same, because the flat file also includes all previous calls in the folder
calls_flat <- rebuild_flat_parquet(calls_results)

# Named list of individual call objects, keyed by their parameter combination.
# Scales automatically with whatever is in calls_results.
calls <- setNames(
  calls_results$call,
  with(
    calls_results,
    paste(pdf, models, tokens, paste0("rep", replicate), sep = "_")
  )
)

# --- Empty-table detection ---------------------------------------------------
# "Successful" API calls (success == TRUE) may still return all-NA tables if
# the token limit was hit mid-response. Identify these cases here.
# A table is "empty" if every value in it is NA.

call_completeness <- calls_flat |>
  group_by(id, table = level_2) |>
  summarise(
    n_non_na = sum(!is.na(value)),
    is_empty = n_non_na == 0,
    .groups = "drop"
  ) |>
  group_by(id) |>
  summarise(
    tables_with_data = sum(!is_empty),
    tables_empty = sum(is_empty),
    .groups = "drop"
  )

call_status <- calls_results |>
  mutate(
    id = as.character(row_number()),
    api_success = map_lgl(call, \(c) isTRUE(c$success))
  ) |>
  left_join(call_completeness, by = "id") |>
  replace_na(list(tables_with_data = 0L, tables_empty = 0L)) |>
  mutate(
    effectively_complete = api_success &
      tables_empty == 0 &
      tables_with_data > 0
  ) |>
  select(
    pdf,
    models,
    tokens,
    replicate,
    api_success,
    tables_with_data,
    tables_empty,
    effectively_complete
  )
```

```{r}
#| label: batch-report

# --- Batch run status report -------------------------------------------------
# Inspect the call list-column directly — before any flattening — to report
# on successes and errors with token counts, broken down by pdf, model, and
# token limit.

batch_run_report <- calls_results |>
  mutate(
    success = map_lgl(call, \(c) isTRUE(c$success)),
    error_msg = map_chr(call, \(c) c$error %||% NA_character_),
    # metadata is a named list of summed token counts for successful calls.
    # Try both naming conventions used across ellmer versions.
    input_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["input_tokens"]] %||% c$metadata[["input"]] %||% NA_real_
        ),
        error = \(e) NA_integer_
      )
    }),
    output_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["output_tokens"]] %||%
            c$metadata[["output"]] %||%
            NA_real_
        ),
        error = \(e) NA_integer_
      )
    })
  )

# Per-replicate detail
batch_run_report |>
  select(
    pdf,
    models,
    tokens,
    replicate,
    success,
    error_msg,
    input_tokens,
    output_tokens
  ) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    replicate = "Rep",
    success = "Success",
    error_msg = "Error",
    input_tokens = "Input Tok",
    output_tokens = "Output Tok"
  ) |>
  color(i = ~ !success, j = ~ success + error_msg, color = "firebrick")
```

```{r}
#| label: batch-report-summary

# Grouped summary: counts and mean token usage by pdf × model × token limit
batch_run_report |>
  group_by(pdf, models, tokens) |>
  summarise(
    n_calls = n(),
    n_success = sum(success),
    n_failed = sum(!success),
    error_types = paste(sort(unique(na.omit(error_msg))), collapse = "; "),
    mean_input_tok = round(mean(input_tokens, na.rm = TRUE)),
    mean_output_tok = round(mean(output_tokens, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    n_calls = "N",
    n_success = "Success",
    n_failed = "Failed",
    error_types = "Error Types",
    mean_input_tok = "Mean Input Tok",
    mean_output_tok = "Mean Output Tok"
  )
```

```{r}
#| label: tbl-all-calls

call_status |> flextable() |> autofit()
```


```{r}
#| label: tbl-all-calls-summary

call_status |>
  group_by(pdf, models, tokens) |>
  summarise(
    n_calls = n(),
    n_complete = sum(effectively_complete),
    n_empty_tables = sum(api_success & !effectively_complete),
    n_failed = sum(!api_success),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit()
```

Now that I come to think of it we do already have canonical versions (even if they were partly LLM-assisted, they've been manually checked.) They're not quite in the perfect format, but we can work with them.

Where does that leave us:

1. Run the extraction (n = 1) for a paper
2. Adapt validation functions to the extracted data (may be easier said than done)
3. Do a second run of one of the papers. Do a pairwise comparison (as above) for consistency. How do we scale to n > 2?
4. Scale...

## eData Validation

We have functions for this, but they're intended to deal with finished data, not fresh from extraction. Still


```{r}
#| label: edata-validation-setup

# One function per table. Each takes the raw LLM-extracted data frame,
# adds dummy values for columns the LLM does not generate, and returns an
# interrogated pointblank agent. Call get_agent_report(agent) to inspect
# any individual table in detail.
#
# Dummied columns are noted inline. Steps that check generated fields
# (SAMPLE_ID, PROTOCOL_ID) will fail — this is expected.
#
# Tables skipped:
#   samples  — LLM schema (sampling_dates, sampling_site_code, …) does not
#              map to the eData Samples table.
#   comments — No pb_validate_*() function exists for this table.

validate_llm_campaign <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(ENTERED_BY = "Dummy", ENTERED_DATE = today()) |>
    pb_validate_campaign(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_references <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      REFERENCE_TYPE = "Journal Article",
      REFERENCE_ID = generate_reference_id(YEAR, AUTHOR, TITLE),
      ACCESS_DATE = today(),
      INSTITUTION = "Dummy", # not extracted by LLM
      DATA_SOURCE = "Primary" # not extracted by LLM
    ) |>
    pb_validate_reference(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_sites <- function(tbl) {
  # SITE_GEOGRAPHIC_FEATURE will likely fail: the LLM returns descriptive
  # text rather than a controlled-vocabulary value.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ALTITUDE_VALUE = 0, # not extracted by LLM
      ALTITUDE_UNIT = "km", # not extracted by LLM
      SITE_COORDINATE_SYSTEM = "Not relevant", # not extracted by LLM
      ENTERED_BY = "Dummy",
      ENTERED_DATE = today()
    ) |>
    pb_validate_sites(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_parameters <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ENTERED_BY = "Dummy",
      MEASURED_TYPE = "Concentration" # not extracted by LLM
    ) |>
    pb_validate_parameters(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_compartments <- function(tbl) {
  # No pb_validate_compartments() in eDataDRF; build the agent inline.
  tbl |>
    rename_with(str_to_upper) |>
    create_agent(label = "eData Compartments Validation") |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT,
      set = environ_compartments_vocabulary()
    ) |>
    col_vals_in_set(
      columns = MEASURED_CATEGORY,
      set = unname(measured_categories_vocabulary())
    ) |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT_SUB,
      set = environ_compartments_sub_vocabulary() |> flatten()
    ) |>
    interrogate(progress = FALSE)
}

validate_llm_biota <- function(tbl) {
  # SITE_CODE, PARAMETER_NAME, ENVIRON_COMPARTMENT, ENVIRON_COMPARTMENT_SUB
  # are not extracted by the LLM and are dummied.
  # SAMPLE_ID will fail the regex check — it is a generated field.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      SITE_CODE = "Dummy",
      PARAMETER_NAME = "Dummy",
      ENVIRON_COMPARTMENT = "Biota",
      ENVIRON_COMPARTMENT_SUB = first(
        environ_compartments_sub_vocabulary()$Biota
      )
    ) |>
    pb_validate_biota(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# TODO: Methods is definitely still a problem.
validate_llm_methods <- function(tbl) {
  # PROTOCOL_ID and CAMPAIGN_NAME are not extracted by the LLM and are dummied.
  # PROTOCOL_ID will fail the regex check — it is a generated field.
  # TODO: PROTOCOL_NAME is failing all time time
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      PROTOCOL_ID = "Dummy",
      CAMPAIGN_NAME = "Dummy"
    ) |>
    pb_validate_methods(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# Thin wrapper: dispatches to each per-table function and returns a named
# list of interrogated agents. NULL tables are silently skipped.
validate_llm_extraction <- function(result) {
  fns <- list(
    campaign = validate_llm_campaign,
    references = validate_llm_references,
    sites = validate_llm_sites,
    parameters = validate_llm_parameters,
    compartments = validate_llm_compartments,
    biota = validate_llm_biota,
    methods = validate_llm_methods
  )
  imap(fns, \(fn, nm) if (!is.null(result[[nm]])) fn(result[[nm]])) |>
    compact()
}

# Summarise failure counts across all tables in a validation list.
# Returns one row per table with step-level and row-level failure counts.
summarise_validation <- function(validation_list) {
  imap(validation_list, \(agent, table_name) {
    agent$validation_set |>
      summarise(
        table = table_name,
        n_steps = n(),
        n_steps_failed = sum(n_failed > 0, na.rm = TRUE),
        rows_failed = sum(n_failed, na.rm = TRUE),
        pct_steps_pass = mean(all_passed, na.rm = TRUE)
      )
  }) |>
    list_rbind()
}

# Run validation across all successful calls and return a long tidy tibble:
# one row per (call × table), with pdf / models / tokens / replicate attached.
#
# For an overall pass rate per call:
#   validate_all_calls(calls_results) |>
#     group_by(pdf, models, tokens, replicate) |>
#     summarise(pct_pass = mean(pct_steps_pass))
validate_all_calls <- function(calls_results) {
  calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      validation_summary = map(call, \(c) {
        validate_llm_extraction(c$result) |>
          summarise_validation()
      })
    ) |>
    select(id, pdf, models, tokens, replicate, validation_summary) |>
    unnest(validation_summary)
}
```


```{r}
#| label: validation-execute

# Keep named agents available for get_agent_report() inspection.
# Index by row-number ID from batch_run_report, e.g. call_validations[["42"]]

# Reuse batch_run_report to identify which calls have a result worth validating.
successful_ids <- batch_run_report$id[batch_run_report$success]
call_validations <- calls[successful_ids] |>
  set_names(successful_ids) |>
  map(\(c) validate_llm_extraction(c$result))

table_summary <- validate_all_calls(calls_results)
```


```{r}
# TODO: Jitter/dodge points
# TODO: Why are some points not coloured?
# TOOD: Replace boxplots with something better
# This approach isn't very good for the sort of steppy data we have.

table_summary |>
  ggplot(aes(x = table, y = pct_steps_pass, fill = models)) +
  geom_point(
    aes(colour = models),
    position = position_jitterdodge(dodge.width = 0.75)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    name = "Proportion of Validation Passed"
  )
```


```{r}
#| label: canonical-data

# --- Load canonical datasets -------------------------------------------------

pdf_zip_map <- c(
  a = "data-raw/1997RitterhoffTraceMetalsIn_AllData_20251030_145522.zip",
  b = "data-raw/2002HellandTransportAndSedimentation_AllData_20251104_075533.zip",
  c = "data-raw/2021WojtasikTemporalIsolationBetween_AllData_20251030_163338.zip",
  d = "data-raw/2022KohlerDistributionPatternOf_AllData_20251217_115443.zip"
)

load_canonical <- function(zip_path) {
  tmp <- tempfile()
  dir.create(tmp)
  unzip(zip_path, exdir = tmp)
  csvs <- list.files(tmp, full.names = TRUE, pattern = "\\.csv$")

  patterns <- c(
    campaign = "Campaign",
    references = "Reference",
    sites = "Sites",
    parameters = "Parameters",
    compartments = "Compartments",
    biota = "Biota",
    methods = "Methods"
  )

  imap(patterns, function(pat, nm) {
    f <- csvs[grepl(pat, csvs, fixed = TRUE)]
    if (length(f) == 0) {
      return(NULL)
    }
    read_csv(f[1], show_col_types = FALSE) |> rename_with(tolower)
  }) |>
    compact()
}

# Load canonical data for all four papers (cached in-memory)
canonical_data <- imap(pdf_zip_map, \(path, id) load_canonical(path))
```


```{r}
#| label: canon-comparison-setup
# --- Variable classification --------------------------------------------------
# var_type:
#   "controlled" — compare with case-insensitive exact match → boolean
#   "continuous" — compute numeric delta (LLM − canonical)
#   "exclude"  — excluded from comparison
#
# canonical_name uses lowercase column names (matching rename_with(tolower) above).
# LLM schema names that differ from canonical names are mapped explicitly
# (e.g. country_iso → country, ocean_iho → area).

variable_meta <- tribble(
  ~table         , ~llm_name                     , ~canonical_name               , ~var_type    ,
  # campaign
  "campaign"     , "campaign_start_date"         , "campaign_start_date"         , "continuous" ,
  "campaign"     , "campaign_end_date"           , "campaign_end_date"           , "continuous" ,
  "campaign"     , "campaign_name"               , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_name_short"         , NA_character_                 , "exclude"    ,
  "campaign"     , "organisation"                , "organisation"                , "exclude"    ,
  "campaign"     , "campaign_comment"            , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_rationale"          , NA_character_                 , "exclude"    ,
  # references
  "references"   , "reference_type"              , "reference_type"              , "controlled" ,
  "references"   , "year"                        , "year"                        , "continuous" ,
  "references"   , "volume"                      , "volume"                      , "continuous" ,
  "references"   , "issue"                       , "issue"                       , "continuous" ,
  "references"   , "author"                      , NA_character_                 , "exclude"    ,
  "references"   , "title"                       , NA_character_                 , "exclude"    ,
  "references"   , "doi"                         , NA_character_                 , "exclude"    ,
  "references"   , "periodical_journal"          , NA_character_                 , "exclude"    ,
  "references"   , "publisher"                   , NA_character_                 , "exclude"    ,
  # sites
  "sites"        , "latitude"                    , "latitude"                    , "continuous" ,
  "sites"        , "longitude"                   , "longitude"                   , "continuous" ,
  "sites"        , "country_iso"                 , "country"                     , "controlled" ,
  "sites"        , "ocean_iho"                   , "area"                        , "controlled" ,
  "sites"        , "site_geographic_feature"     , "site_geographic_feature"     , "controlled" ,
  "sites"        , "site_geographic_feature_sub" , "site_geographic_feature_sub" , "controlled" ,
  "sites"        , "site_code"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_name"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_comment"                , NA_character_                 , "exclude"    ,
  # parameters
  "parameters"   , "parameter_type"              , "parameter_type"              , "controlled" ,
  "parameters"   , "cas_rn"                      , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_name"              , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_comment"           , NA_character_                 , "exclude"    ,
  # compartments
  "compartments" , "environ_compartment"         , "environ_compartment"         , "controlled" ,
  "compartments" , "environ_compartment_sub"     , "environ_compartment_sub"     , "controlled" ,
  "compartments" , "measured_category"           , "measured_category"           , "controlled" ,
  # biota
  "biota"        , "species_group"               , "species_group"               , "controlled" ,
  "biota"        , "sample_tissue"               , "sample_tissue"               , "controlled" ,
  "biota"        , "sample_species_lifestage"    , "sample_species_lifestage"    , "controlled" ,
  "biota"        , "sample_species_gender"       , "sample_species_gender"       , "controlled" ,
  "biota"        , "sample_species"              , NA_character_                 , "exclude"    ,
  "biota"        , "sample_id"                   , NA_character_                 , "exclude"    ,
  # methods
  "methods"      , "protocol_category"           , "protocol_category"           , "controlled" ,
  "methods"      , "protocol_name"               , "protocol_name"               , "controlled" ,
  "methods"      , "protocol_comment"            , NA_character_                 , "exclude"    ,
)

# Row-matching key columns per table (NULL = single-row, match directly)
# TODO: Site matching via rounded lat/lon is coarse (~111 m tolerance per 0.001°).
#       Improve with st_nearest_feature() or site_name fuzzy matching.
table_keys <- list(
  campaign = NULL,
  references = NULL,
  sites = c("latitude", "longitude"),
  parameters = "parameter_name",
  compartments = c("environ_compartment", "environ_compartment_sub"),
  biota = c("sample_species", "sample_tissue"),
  methods = c("protocol_category", "protocol_name")
)
```

I was going to try and do some more sophisticated matching here, but it looks like everything we get from the call is a string anyway. So I think it's best to just compare strings. ==, True or False, sum.

```{r}
#| label: canon-comparison-test

# so, pick a call
calls_flat_1 <- calls_flat |> filter(id == 1)

calls_flat_1
canonical_1 <- canonical_data$a

# pick a table
calls_flat_1_campaign <- filter(calls_flat_1, level_2 == "campaign") |>
  select(name, value, type) # fuck, types are wrong

canonical_1_campaign <- canonical_1$campaign |>
  pivot_longer(cols = everything(), values_transform = as.character)


# TODO: Actually, we're overthinking
joined_tables <- left_join(
  calls_flat_1_campaign,
  canonical_1_campaign,
  by = "name",
  suffix = c(".call", ".canon"),
) |>
  left_join(
    variable_meta |> filter(table == "campaign"),
    by = c(name = "canonical_name")
  )

# actually, just run check as characters
valid_comparisons <- joined_tables |>
  filter(!is.na(var_type) & var_type != "exclude") |>
  mutate(match = identical(value.canon, value.call, single.NA = FALSE))
```


```{r}
#| label: canon-comparison-all

# Scale string comparison across all calls × all tables.
#
# For each (call_id, table) pair:
#   1. Align LLM field names to canonical field names via variable_meta.
#   2. Pool all values across rows (ignoring row identity).
#   3. Per field: any LLM value == any canonical value → match.
# Produces one row per (call × table) with n_match, n_total, prop_match.

canonical_comparison <- calls_flat |>
  distinct(id, pdf, models, tokens, replicate) |>
  crossing(
    table = variable_meta |>
      filter(!is.na(var_type), var_type != "exclude") |>
      distinct(table) |>
      pull(table)
  ) |>
  mutate(
    result = pmap(list(id, pdf, table), \(call_id, pdf_id, tbl) {
      canonical_tbl <- canonical_data[[pdf_id]][[tbl]]
      if (is.null(canonical_tbl)) {
        return(NULL)
      }

      meta <- variable_meta |>
        filter(
          .data$table == tbl,
          !is.na(.data$var_type),
          .data$var_type != "exclude"
        )

      llm_vals <- calls_flat |>
        filter(id == call_id, level_2 == tbl) |>
        select(name, value) |>
        inner_join(
          meta |> select(llm_name, canonical_name),
          by = c(name = "llm_name")
        ) |>
        select(canonical_name, value_call = value)

      canon_vals <- canonical_tbl |>
        pivot_longer(cols = everything(), values_transform = as.character) |>
        inner_join(
          meta |> select(canonical_name),
          by = c(name = "canonical_name")
        ) |>
        select(canonical_name = name, value_canon = value)

      if (nrow(llm_vals) == 0 || nrow(canon_vals) == 0) {
        return(NULL)
      }

      # For each field: does any LLM value match any canonical value?
      llm_vals |>
        left_join(
          canon_vals,
          by = "canonical_name",
          relationship = "many-to-many"
        ) |>
        group_by(canonical_name) |>
        summarise(
          match = any(value_call == value_canon, na.rm = TRUE),
          .groups = "drop"
        ) |>
        summarise(
          n_match = sum(match, na.rm = TRUE),
          n_total = n(),
          prop_match = n_match / n_total
        )
    })
  ) |>
  filter(!map_lgl(result, is.null)) |>
  unnest(result)

canonical_comparison
```

```{r}
#| label: canon-comparison-all-summary

# what's the best way to summarise this?
canonical_comparison_summarised <- canonical_comparison |>
  group_by(models, tokens) |>
  reframe(mean_prop_match = mean(prop_match), n_replicates = max(replicate))

# our questions are:
# which models are most accurate?
# which tables cause the most problems?
# does token count affect anything?

# for a standard "look how good my LLM is" comparison, we'd just do a bar chart...

canonical_comparison_summarised |>
  ggplot(aes(x = models, y = mean_prop_match, fill = tokens)) +
  geom_col()

# realistically, we should at the very least do summary stats
canonical_comparison |>
  ggplot(aes(x = models, y = prop_match, colour = table)) +
  geom_boxplot()

# again, we have the issue that points overplot. better with boxplots (sort of)
```

```{r}
#| label: compare-canonical-comments

# Screening comments are a special case: there is no canonical ground truth
# to compare against, so we assess inter-rater reliability between
# models/replicates using Cohen's / Fleiss' kappa.
#
# Weighted kappa is used throughout: it penalises larger disagreements more
# (e.g. 1 vs 5) than small ones (1 vs 2), which is appropriate for 1–5
# ordinal scores.

score_vars <- c(
  "paper_relevance",
  "paper_reliability",
  "paper_data_source",
  "paper_data_available",
  "extraction_assessement"
)

# Extract integer scores from the "Score: N: text" format
scores_long <- calls_flat |>
  filter(level_2 == "comments", name %in% score_vars) |>
  select(id, pdf, models, tokens, replicate, name, value) |>
  mutate(score = as.integer(str_extract(value, "(?<=Score:\\s)\\d"))) |>
  select(-value)


# --- Within-model kappa (consistency across replicates) ----------------------
# For each (model, score_variable): matrix rows = pdfs, cols = replicates.
# Multiple raters → Fleiss' kappa.

within_model_kappa <- scores_long |>
  group_by(models, name) |>
  group_map(function(df, key) {
    mat <- df |>
      select(pdf, replicate, score) |>
      pivot_wider(names_from = replicate, values_from = score) |>
      select(-pdf) |>
      as.matrix()

    # Drop all-NA columns (failed/empty replicates)
    mat <- mat[, colSums(!is.na(mat)) > 0, drop = FALSE]
    if (ncol(mat) < 2 || nrow(mat) < 2) {
      return(tibble())
    }

    res <- tryCatch(kappam.fleiss(mat), error = function(e) NULL)
    if (is.null(res)) {
      return(tibble())
    }

    tibble(
      models = key$models,
      score_var = key$name,
      n_subjects = nrow(mat),
      n_raters = ncol(mat),
      kappa = round(res$value, 3),
      z = round(res$statistic, 2),
      p_value = round(res$p.value, 4),
      comparison = "within_model (replicates)"
    )
  }) |>
  list_rbind()


# --- Cross-model kappa (agreement between models) ----------------------------
# Average scores across replicates per (pdf, model) first, then compare models.
# 2 models → weighted kappa2; >2 models → Fleiss kappa.

cross_model_kappa <- scores_long |>
  group_by(pdf, models, name) |>
  summarise(mean_score = round(mean(score, na.rm = TRUE)), .groups = "drop") |>
  group_by(name) |>
  group_map(function(df, key) {
    mat <- df |>
      select(pdf, models, mean_score) |>
      pivot_wider(names_from = models, values_from = mean_score) |>
      select(-pdf) |>
      as.matrix()

    mat <- mat[, colSums(!is.na(mat)) > 0, drop = FALSE]
    if (ncol(mat) < 2 || nrow(mat) < 2) {
      return(tibble())
    }

    res <- tryCatch(
      if (ncol(mat) == 2) {
        kappa2(mat, weight = "equal")
      } else {
        kappam.fleiss(mat)
      },
      error = function(e) NULL
    )
    if (is.null(res)) {
      return(tibble())
    }

    tibble(
      models = "all",
      score_var = key$name,
      n_subjects = nrow(mat),
      n_raters = ncol(mat),
      kappa = round(res$value, 3),
      z = round(res$statistic, 2),
      p_value = round(res$p.value, 4),
      comparison = "cross_model"
    )
  }) |>
  list_rbind()
```


```{r}
#| label:  tbl-combined-results

# Combined results table
bind_rows(within_model_kappa, cross_model_kappa) |>
  arrange(comparison, score_var, models) |>
  flextable() |>
  autofit()
```



```{r}
#| label: call-comparison
calls_flat

# KET is right, pairwise might get messy quickly

# campaign and reference are easy.

# almost certainly there are better options than identical()
compare_campaign <- calls_flat |>
  filter(level_2 == "campaign") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# no issues with references
compare_references <- calls_flat |>
  filter(level_2 == "references") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# sites...
# issue: different calls return variables in different orders!
# we can't automatically match sites to each other unless we assume that the lat/long are always equal (or, if we have a canonical point of comparison, I guess)
compare_sites <- calls_flat |>
  filter(level_2 == "sites") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# we can presumably compare distance with sf

# for example
compare_sites |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric))

# how do we check if each value has an equivalent?
# I guess we just group and ignore idx
# we may have to call st_distance for g
distances <- calls_flat |>
  filter(level_2 == "sites") |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric)) |>
  group_by(id, row_idx) |>
  pivot_wider(names_from = name, values_from = value) |>
  st_as_sf(coords = c("latitude", "longitude"), na.fail = FALSE)


# pair sites and check (great circle) distance?

x <- distances[1:3, 6]
y <- distances[4:6, 6]

# I guess we make a distance matrix, which should have one value of approximately 0 and be... orthogonal?
# I believe these distances are in km, would be good to check
st_distance(x, y)

# parameters should be easier, they have more-or-less canonical names
compare_parameters <- calls_flat |>
  filter(level_2 == "parameters") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# compartments - easy
compare_compartments <- calls_flat |>
  filter(level_2 == "compartments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# biota
# common or linnean name is acceptable as long as they agree
compare_biota <- calls_flat |>
  filter(level_2 == "biota") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# methods
# again, pairing may be an issue here
compare_methods <- calls_flat |>
  filter(level_2 == "methods") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# samples
# might not do the evaluation on this one, since we don't strictly speaking use it
compare_samples <- calls_flat |>
  filter(level_2 == "samples") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# comments are pleasantly easy
compare_comments <- calls_flat |>
  filter(level_2 == "comments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call") |>
  mutate(
    score_1 = str_extract(call1, "(?!Score\\: )[1-5]"),
    score_2 = str_extract(call2, "(?!Score\\: )[1-5]"),
    .before = call1
  )
```