---
title: "LLM Evaluation"
subtitle: "Comparing the quality, speed, and cost of LLM extraction"
knitr:
  opts_chunk:
    collapse: true
    dpi: 300                         # belt-and-suspenders for knitr devices
    dev: "ragg_png"                  # use ragg for better font rendering
execute:
  error: true
  cache: true
format: 
  html:
    code-fold: true
  pdf: 
    echo: false
    warning: false
  docx:  
    fig-width: 16
    fig-dpi: 300                     # high res
    fig-format: png                  # png is fine; svg not well supported in docx
    link-citations: true             # TODO: Not working... why?
    link-bibliography: true

filters:
  - flextable-qmd                    # needed, or the flextables render crazy 
  - at: post-render
    path: ../../_extensions/flextable-qmd/unwrap-float.lua
---

```{r}
#| label: setup
#| echo: false
#| include: false

library(STOPeData)
library(ellmer)
library(tidyverse, mask.ok = c("filter", "lag"))
library(flextable)
library(arrow)
library(sf)
library(eDataDRF)
library(pointblank)
library(irr)
library(here)
library(patchwork)

devtools::load_all() # remove once STOPeData is rebuilt

set_flextable_defaults(
  font.size = 10,
  # font.family = 'Aptos',
  digits = 2,
  theme_fun = theme_vanilla,
  padding = 1
)

# Create a location to save LLM API calls, if it doesn't exist.
LLM_CACHE_DIR <- here("data-raw/llm_cache/calls")
dir.create(LLM_CACHE_DIR, recursive = TRUE, showWarnings = FALSE)

i_am("README.md")
```

# Introduction

The `ellmer` package includes functions that connect to a wide variety of local and remote APIs from a variety of providers. Put simply, we have a great many options. Which are better, and which are worse? We might assume that more expensive models are more effective at data extraction, and that larger and more complex files are more difficult - and require more tokens - than simpler models when extracting data. But can we prove any of these assumptions?

Token size defines the memory space available to the LLM, and the cost of extraction. 

# Methods

```{mermaid}
%%| label: workflow-diagram
%%| fig-cap: "Overview of the LLM evaluation pipeline."
flowchart TD
    pdfs["PDFs tibble<br>4 papers"]         --> grid
    models["Models tibble<br>3 providers"]  --> grid
    tok["Token limits<br>3k / 6k / 9k<br>n replicates"] --> grid

    grid["Experiment grid<br>run_extraction_batch"]  --> cache

    cache{"RDS cache hit?"}
    cache -- Yes --> load["Load cached RDS"]
    cache -- No  --> api["LLM API call<br>extract_pdf_with_llm"]

    api -- success --> save["Save RDS"]
    api -- failure --> err["Record error<br>HTTP code + message"]

    save  --> results
    load  --> results
    err   --> results

    results["calls_results<br>call list-column"]

    results --> flat["rebuild_flat_parquet<br>Successful calls only"]
    flat    --> parquet[("calls_flat parquet<br>long / flat format")]

    results --> report["batch_run_report<br>success · error · tokens<br>by pdf · model · limit"]

    parquet --> status["call_status<br>success · empty-table<br>effectiveness flags"]
    parquet --> validate["validate_all_calls<br>pointblank agents<br>per table"]
    parquet --> canon["Canonical comparison<br>vs. human-extracted data<br>controlled + continuous vars"]
    parquet --> kappa["Kappa statistics<br>inter-rater reliability<br>within- and cross-model"]
```

## Study Materials

We selected four studies [@tbl-pdfs] that were previously identified as high-quality (score >= 4), covering a variety of time periods, formats, topics, and methods of presenting data.  

```{r}
#| label: tbl-pdfs
#| tbl-cap: "Study materials (portable document format files) selected for LLM evaluation. OCR (Optical Character Recognition) refers to cases where the paper used has been scanned from a printed copy and automatically converted to text. Depending on the quality of this process, information may be lost or corrupted; we would expect such files to be more difficult to extract data from."

pdfs <- tibble(
  id = c("a", "b", "c", "d"),
  path = c(
    here("data-raw/1997RitterhoffTraceMetalsIn_MS.pdf"),
    here("data-raw/2002HellandTransportAndSedimentation_MS.pdf"),
    here("data-raw/2021WojtasikTemporalIsolationBetween_MS.pdf"),
    here("data-raw/2022KohlerDistributionPatternOf_MS.pdf")
  ),
  title = c(
    "Trace metals in field samples of zooplankton from the Fram Strait and the Greenland Sea",
    "Temporal isolation between two strongly differentiated stocks of the Greenland halibut (Reinhardtius hippoglossoides, Walbaum, 1792) from the Western Barents Sea",
    "Transport and sedimentation of Cu in a microtidal estuary, SE Norway",
    "Distribution pattern of mercury in northern Barents Sea and Eurasian Basin surface sediment"
  ),
  year = c(1997L, 2002L, 2021L, 2022L),
  comment = c(
    "Well-OCR'd paper with summary statistics of trace metal concentrations reported across wide range zooplankton taxa. Sampling sites reported in low-quality map. Complex, uneven data structure due to random nature of zooplankton sampling. Methodology reported in moderate detail.",
    "Native PDF.  Sites reported in map. Cu concentrations (and flux) reported from a variety of sediment traps at different timepoints. Methodology reported in moderate detail.",
    "Native PDF. Study primarily concerns evolution but includes heavy metal analysis. Sites reported as linear trawls with start/end coordinates. Raw Cu concentrations given for each individual, reported as table. Methodology reported in high detail.",
    "Native PDF. Sampling sites and Cu concentrations only reported as summaries/graphics in paper, but raw data (Cu concentrations, site coordinates, depth) provided on request and prepended to file."
  ),
  # Fill in after reviewing each PDF
  num_images = c(1, 4, 8, 5),
  num_tables = c(7, 2, 8, 2)
) |>
  mutate(
    info = map(path, pdftools::pdf_info),
    file_size_kb = map_dbl(path, \(p) file.size(p) / 1024),
    num_pages = map_int(info, \(x) x$pages)
  ) |>
  select(-info)

pdfs |>
  select(-path) |>
  flextable() |>
  autofit() |>
  colformat_double(j = c("file_size_kb")) |>
  colformat_int(j = c("year"), big.mark = "") |>
  set_header_labels(
    id = "ID",
    title = "Title",
    year = "Year",
    comment = "Comment",
    num_images = "n Images",
    num_tables = "n Tables",
    file_size_kb = "File Size (kb)",
    num_pages = "n Pages"
  )
```

## Large Language Models

The R `ellmer` package includes API integrations for a variety of Large Language Models from various providers. We selected general, "middle-of-the-road" multimodal models from three of the most prominent providers - Anthropic, Google Gemini, and OpenAI, as well as immediately higher and lower tier models. In addition, we wished to validate the performance of local and/or open-source LLMs. This proved technically difficult, and with the computational resources available we were unable to run models locally at acceptable levels of performance (models could technically be run, but lacked the available memory to load PDF files). As a compromise, we also include the open model Gemma. Gemma is developed by Google (and thus accessible via their API to be run on their servers), but can also be loaded and run locally using services such as Ollama (which supports structured data extraction) and LMStudio and accessed locally with ellmer via their APIs. 

Our selected models are summarised in [@tbl-models]. Details are sourced from providers' model overview pages.

Sources for model comparisons: 
https://platform.claude.com/docs/en/about-claude/models/overview, 
https://developers.openai.com/api/docs/models/compare, 
https://ai.google.dev/gemini-api/docs/models, 
https://ai.google.dev/gemini-api/docs/pricing

NB: We can't report a direct cost for Gemma, because it's free (but rate-limited) from Google, or "free" if you run it locally. 

```{r}
#| label: tbl-models
#| tbl-cap: "Summary of LLM models included in the evaluation. Tok: tokens, a unit of LLM knowledge; Mtok: million tokens."

# TODO: Fill out list

models_tbl <- tribble(
  ~provider                          , ~model_label                         , ~model_id                ,
  ~context_window_k                  , ~max_output_k                        ,
  ~cost_input_mtok                   , ~cost_output_mtok                    ,
  ~knowledge_cutoff                  , ~type                                ,
  # ---- Anthropic ---------------------------------------------------------------
  "Anthropic"                        , "Claude Haiku 4.5"                   , "claude-haiku-4-5"       ,
  "200"                              , "64"                                 ,
  "1"                                , "5"                                  ,
  "Feb 2025"                         , "Closed"                             ,
  #
  "Anthropic"                        , "Claude Sonnet 4.6"                  , "claude-sonnet-4-6"      ,
  "1000"                             , "128"                                ,
  "3"                                , "15"                                 ,
  "Aug 2025"                         , "Closed"                             ,
  #
  "Anthropic"                        , "Claude Opus 4.8"                    , "claude-opus-4-8"        ,
  "1000"                             , "128"                                ,
  "5"                                , "25"                                 ,
  "Jan 2026"                         , "Closed"                             ,
  # ---- OpenAI ------------------------------------------------------------------
  "OpenAI"                           , "GPT-5.4 Mini"                       , "gpt-5.4-mini"           ,
  "400"                              , "128"                                ,
  "0.75"                             , "4.50"                               ,
  "Aug 2025"                         , "Closed"                             ,
  #
  "OpenAI"                           , "GPT-5.4"                            , "gpt-5.4"                ,
  "1050"                             , "128"                                ,
  "2.5"                              , "15.0"                               ,
  "Aug 2025"                         , "Closed"                             ,
  #
  "OpenAI"                           , "GPT-5.5"                            , "gpt-5.5"                ,
  "1050"                             , "128"                                ,
  "cost_input_mtok"                  , "30.0"                               ,
  "Dec 2025"                         , "Closed"                             ,
  # ---- Google Gemini -----------------------------------------------------------
  "Google Gemini"                    , "Gemini 2.5 Pro"                     , "gemini-2.5-pro"         ,
  "1050"                             , "65"                                 ,
  "1.25 (<= 200k tokens), then 2.5"  , "10.00 (<= 200k tokens), then 15.00" ,
  "Jan 2025"                         , "Closed"                             ,
  #
  "Google Gemini"                    , "Gemini 3.1 Pro Preview"             , "gemini-3.1-pro-preview" ,
  "context_window_k"                 , "max_output_k"                       ,
  "2.00 (<= 200k tokens), then 4.00" , "12.00 (<= 200k tokens), then 18.00" ,
  "knowledge_cutoff"                 , "Closed"                             ,
  #
  "Google Gemini"                    , "Gemini 3.5 Flash"                   , "gemini-3.5-flash"       ,
  "1000"                             , "65"                                 ,
  "1.50"                             , "9.00"                               ,
  "Jan 2025"                         , "Closed"                             ,
  # ---- Google Gemma (open-weights) ---------------------------------------------
  "Google Gemini"                    , "Gemma 4 31B"                        , "gemma-4-31b-it"         ,
  "262"                              , "8k"                                 ,
  "Free"                             , "Free"                               ,
  "Jan 2025"                         , "Open"
)
# Named vector for use in experimental_setup (label -> API id)
models <- setNames(models_tbl$model_id, models_tbl$model_label)


models_tbl |>
  flextable() |>
  autofit() |>
  set_header_labels(
    provider = "Provider",
    model_label = "Model",
    model_id = "Model ID",
    context_window_k = "Context Window (Ktok)",
    max_output_k = "Max Output (Ktok)",
    cost_input_mtok = "Input Cost ($/Mtok)",
    cost_output_mtok = "Output Cost ($/Mtok)",
    knowledge_cutoff = "Knowledge Cutoff",
    training_cutoff = "Training Cutoff",
    type = "Open/Closed"
  )
```

## API Calls

Provider APIs are called using the existing API call functionality developed for eData (`STOPeData::extract_pdf_with_llm()`). Each call returns a list object containing extracted data and call metadata (cost, tokens, etc.), which is then saved as as an `.rds` file. 

```{r}
#| label: API-calls-setup
#| code-fold: true

# Recursively walk the list, collecting path-as-vector as we go.
# Each branch terminates at: (1) a data frame, (2) an atomic scalar.
flatten_to_rows <- function(x, path = character()) {
  if (is.data.frame(x)) {
    # --- TIBBLE BRANCH ---
    # Capture column types before pivoting (pivot coerces everything to one type)
    col_types <- sapply(x, \(col) paste(class(col), collapse = "/"))

    x |>
      mutate(.row_idx = row_number()) |>
      pivot_longer(
        cols = -.row_idx,
        names_to = "name",
        values_to = "value",
        values_transform = as.character # coerce all cells → chr for uniform value col
      ) |>
      mutate(
        level_path = list(path), # store the full path as a list-col
        row_idx = .row_idx,
        type = col_types[name] # look up original type by col name
      ) |>
      select(level_path, name, row_idx, value, type)
  } else if (is.list(x)) {
    # --- LIST BRANCH: recurse into each named element ---
    imap(x, \(item, nm) flatten_to_rows(item, c(path, nm))) |>
      list_rbind()
  } else {
    # --- SCALAR BRANCH ---
    # The last segment of path IS the name; everything before is the path.
    tibble(
      level_path = list(head(path, -1)),
      name = tail(path, 1),
      row_idx = NA_integer_,
      value = as.character(x),
      type = paste(class(x), collapse = "/")
    )
  }
}

# Map provider display names → the ellmer chat function and API key env var.
# model_provider is the string extract_pdf_with_llm() checks internally.
provider_config <- tribble(
  ~provider       , ~model_provider , ~env_var            , ~chat_fn             ,
  "Anthropic"     , "Anthropic"     , "ANTHROPIC_API_KEY" , "chat_anthropic"     ,
  "OpenAI"        , "OpenAI"        , "OPENAI_API_KEY"    , "chat_openai"        ,
  "Google Gemini" , "Google"        , "GOOGLE_API_KEY"    , "chat_google_gemini"
)

# Run a single extraction. Looks up the API key from the env var at call time.
run_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn
) {
  extract_pdf_with_llm(
    pdf_path = pdf_path,
    model_provider = model_provider,
    model_name = model_id,
    env_var = env_var,
    chat_fn = chat_fn,
    api_key = Sys.getenv(env_var),
    extraction_prompt = create_extraction_prompt(),
    extraction_schema = create_extraction_schema(),
    max_tokens = as.integer(max_tokens)
  )
}

# Unpack the level_path list-col into level_1 / level_2 character columns.
flatten_call <- function(call_result) {
  flatten_to_rows(call_result) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)
}

# --- Per-call RDS caching ----------------------------------------------------

# Build a deterministic RDS path from the call parameters.
call_rds_path <- function(
  pdf,
  model_id,
  tokens,
  replicate,
  cache_dir = LLM_CACHE_DIR
) {
  # Replace path-unsafe characters in model_id (e.g. "/")
  safe_model <- gsub("/", "-", model_id)
  file.path(
    cache_dir,
    glue("{pdf}_{safe_model}_{tokens}_rep{replicate}.rds")
  )
}

# Run a single extraction, loading from RDS cache if already done.
run_or_load_extraction <- function(
  pdf_path,
  model_id,
  max_tokens,
  model_provider,
  env_var,
  chat_fn,
  pdf,
  replicate
) {
  rds_path <- call_rds_path(pdf, model_id, max_tokens, replicate)

  if (file.exists(rds_path)) {
    message(glue("Cache hit — loading: {basename(rds_path)}"))
    result <- readRDS(rds_path)
  } else {
    message(glue("Cache miss — running API call: {basename(rds_path)}"))
    result <- run_extraction(
      pdf_path,
      model_id,
      max_tokens,
      model_provider,
      env_var,
      chat_fn
    )
    saveRDS(result, rds_path)
  }
  result
}


# Build an experiment grid and run/load all extractions.
# Each call is cached individually as an RDS keyed on its parameters.
# Returns a tibble with a `call` list-column holding list(result, metadata, success).
#
# pdf_ids:      character vector of PDF ids from the pdfs tibble (e.g. c("a", "b"))
# model_ids:    character vector of model IDs from models_tbl$model_id
# token_counts: integer vector of max_tokens values to test
# n_replicates: how many times to repeat each combination
run_extraction_batch <- function(
  pdf_ids,
  model_ids,
  token_counts,
  n_replicates
) {
  expand.grid(
    pdf = pdf_ids,
    models = model_ids,
    tokens = token_counts,
    replicate = seq_len(n_replicates),
    stringsAsFactors = FALSE
  ) |>
    as_tibble() |>
    left_join(pdfs |> select(id, path), by = c("pdf" = "id")) |>
    left_join(
      models_tbl |>
        select(model_id, provider) |>
        left_join(provider_config, by = "provider"),
      by = c("models" = "model_id")
    ) |>
    mutate(
      call = pmap(
        list(
          pdf_path = path,
          model_id = models,
          max_tokens = tokens,
          model_provider = model_provider,
          env_var = env_var,
          chat_fn = chat_fn,
          pdf = pdf,
          replicate = replicate
        ),
        run_or_load_extraction
      )
    )
}

# Rebuild the flat parquet from all per-call RDS files present in the cache.
# Called every time — cheap, since it just reads already-saved RDS files.
rebuild_flat_parquet <- function(
  calls_results,
  parquet_path = here("data-raw/llm_cache/llm_eval_output_flat.parquet")
) {
  calls_flat <- calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      id = as.character(row_number()),
      flat = map(call, flatten_call)
    ) |>
    select(id, pdf, models, tokens, replicate, flat) |>
    unnest(flat)

  write_parquet(calls_flat, parquet_path)
  calls_flat
}
```

For each manuscript and model multiple calls were made, with token count was set to 3,000, 6,000 and 9,000 to determine if call success, effectiveness, and quality varied with this. Each combination of call, model, token count, and pdf was made 5 times, to provide a measure of inter- and intra-call variation.

```{r}
#| label: tbl-experimental-design
#| tbl-cap: "Experimental design: combinations of PDFs, models, and token limits."

tibble(
  Parameter = c(
    "PDF IDs",
    "Model IDs",
    "Token Limits",
    "Replicates per Combination"
  ),
  Values = c(
    paste(c("a", "b", "c", "d"), collapse = ", "),
    paste(c("claude-haiku-4-5", "claude-sonnet-4-6"), collapse = ", "),
    paste(c(3000, 6000, 9000), collapse = ", "),
    "5"
  )
) |>
  flextable() |>
  autofit() |>
  set_header_labels(Parameter = "Parameter", Values = "Values")
```

```{r}
#| label: api-call

# RUNME

# this is where we set the actual experimental parameters
calls_results <- run_extraction_batch(
  pdf_ids = c("a", "b", "c"),
  model_ids = c(
    "claude-haiku-4-5",
    "claude-sonnet-4-6",
    "gpt-5.4-mini",
    "gpt-5.4",
    "gemini-3.5-flash"
  ),
  token_counts = c(3000, 6000, 9000),
  n_replicates = 3
) |>
  mutate(id = row_number(), .before = everything())

# nb: the number of IDs in calls_results and calls_flats may not be the same, because the flat file also includes all previous calls in the folder
calls_flat <- rebuild_flat_parquet(calls_results)

# Named list of individual call objects, keyed by their parameter combination.
# Scales automatically with whatever is in calls_results.
calls <- setNames(
  calls_results$call,
  with(
    calls_results,
    paste(pdf, models, tokens, paste0("rep", replicate), sep = "_")
  )
)

# --- Empty-table detection ---------------------------------------------------
# "Successful" API calls (success == TRUE) may still return all-NA tables if
# the token limit was hit mid-response. Identify these cases here.
# A table is "empty" if every value in it is NA.

call_completeness <- calls_flat |>
  group_by(id, table = level_2) |>
  summarise(
    n_non_na = sum(!is.na(value)),
    is_empty = n_non_na == 0,
    .groups = "drop"
  ) |>
  group_by(id) |>
  summarise(
    tables_with_data = sum(!is_empty),
    tables_empty = sum(is_empty),
    .groups = "drop"
  )

call_status <- calls_results |>
  mutate(
    id = as.character(row_number()),
    api_success = map_lgl(call, \(c) isTRUE(c$success))
  ) |>
  left_join(call_completeness, by = "id") |>
  replace_na(list(tables_with_data = 0L, tables_empty = 0L)) |>
  mutate(
    effectively_complete = api_success &
      tables_empty == 0 &
      tables_with_data > 0
  ) |>
  select(
    id,
    pdf,
    models,
    tokens,
    replicate,
    api_success,
    tables_with_data,
    tables_empty,
    effectively_complete
  )
```

## Evaluation

Model effectiveness and efficiency was evaluated by considering cost, proportion of data passing evaluation, and proportion of data matching canonical (human-curated) extracted datasets. 

### Cost

API calls return expended input tokens (used to ingest the data, schema, and prompt) and output tokens (used to create the response). Costs per token are reported for most closed models (see [@tbl-...]). Overall cost is therefore calculated as follows:


$$C = t_i c_i + t_o c_o$$

In the case of the open model Gemma 4 (accessed via the Google Gemini API), usage is free, but rate-limited. 

### eData Validation 

Extracted data are validated against the validation functions described in `eDataDRF`, which check the data against a number of rules per table. These functions are intended to validate final data from the eData application after LLM extraction and subsequent human verification, and therefore some variables were "dummied" to avoid irrelevant failures to validate. The number of validation checks passed per table divided by total number of validation checks was used as a simple measure of LLMs' ability to return data consistent with the extraction schema. 


```{r}
#| label: edata-validation-setup

# One function per table. Each takes the raw LLM-extracted data frame,
# adds dummy values for columns the LLM does not generate, and returns an
# interrogated pointblank agent. Call get_agent_report(agent) to inspect
# any individual table in detail.
#
# Dummied columns are noted inline. Steps that check generated fields
# (SAMPLE_ID, PROTOCOL_ID) will fail — this is expected.
#
# Tables skipped:
#   samples  — LLM schema (sampling_dates, sampling_site_code, …) does not
#              map to the eData Samples table.
#   comments — No pb_validate_*() function exists for this table.

validate_llm_campaign <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(ENTERED_BY = "Dummy", ENTERED_DATE = today()) |>
    pb_validate_campaign(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_references <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      REFERENCE_TYPE = "Journal Article",
      REFERENCE_ID = generate_reference_id(YEAR, AUTHOR, TITLE),
      ACCESS_DATE = today(),
      INSTITUTION = "Dummy", # not extracted by LLM
      DATA_SOURCE = "Primary" # not extracted by LLM
    ) |>
    pb_validate_reference(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_sites <- function(tbl) {
  # SITE_GEOGRAPHIC_FEATURE will likely fail: the LLM returns descriptive
  # text rather than a controlled-vocabulary value.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ALTITUDE_VALUE = 0, # not extracted by LLM
      ALTITUDE_UNIT = "km", # not extracted by LLM
      SITE_COORDINATE_SYSTEM = "Not relevant", # not extracted by LLM
      ENTERED_BY = "Dummy",
      ENTERED_DATE = today()
    ) |>
    pb_validate_sites(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_parameters <- function(tbl) {
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      ENTERED_BY = "Dummy",
      MEASURED_TYPE = "Concentration" # not extracted by LLM
    ) |>
    pb_validate_parameters(agent = TRUE) |>
    interrogate(progress = FALSE)
}

validate_llm_compartments <- function(tbl) {
  # No pb_validate_compartments() in eDataDRF; build the agent inline.
  tbl |>
    rename_with(str_to_upper) |>
    create_agent(label = "eData Compartments Validation") |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT,
      set = environ_compartments_vocabulary()
    ) |>
    col_vals_in_set(
      columns = MEASURED_CATEGORY,
      set = unname(measured_categories_vocabulary())
    ) |>
    col_vals_in_set(
      columns = ENVIRON_COMPARTMENT_SUB,
      set = environ_compartments_sub_vocabulary() |> flatten()
    ) |>
    interrogate(progress = FALSE)
}

validate_llm_biota <- function(tbl) {
  # SITE_CODE, PARAMETER_NAME, ENVIRON_COMPARTMENT, ENVIRON_COMPARTMENT_SUB
  # are not extracted by the LLM and are dummied.
  # SAMPLE_ID will fail the regex check — it is a generated field.
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      SITE_CODE = "Dummy",
      PARAMETER_NAME = "Dummy",
      ENVIRON_COMPARTMENT = "Biota",
      ENVIRON_COMPARTMENT_SUB = first(
        environ_compartments_sub_vocabulary()$Biota
      )
    ) |>
    pb_validate_biota(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# TODO: Methods is definitely still a problem.
validate_llm_methods <- function(tbl) {
  # PROTOCOL_ID and CAMPAIGN_NAME are not extracted by the LLM and are dummied.
  # PROTOCOL_ID will fail the regex check — it is a generated field.
  # TODO: PROTOCOL_NAME is failing all time time
  tbl |>
    rename_with(str_to_upper) |>
    mutate(
      PROTOCOL_ID = "Dummy",
      CAMPAIGN_NAME = "Dummy"
    ) |>
    pb_validate_methods(agent = TRUE) |>
    interrogate(progress = FALSE)
}

# Thin wrapper: dispatches to each per-table function and returns a named
# list of interrogated agents. NULL tables are silently skipped.
validate_llm_extraction <- function(result) {
  fns <- list(
    campaign = validate_llm_campaign,
    references = validate_llm_references,
    sites = validate_llm_sites,
    parameters = validate_llm_parameters,
    compartments = validate_llm_compartments,
    biota = validate_llm_biota,
    methods = validate_llm_methods
  )
  imap(fns, \(fn, nm) {
    tbl <- result[[nm]]
    if (is.null(tbl)) {
      return(NULL)
    }
    if (nrow(tbl) == 0) {
      return(list(.empty = TRUE))
    }
    fn(tbl)
  }) |>
    compact()
}

# Summarise failure counts across all tables in a validation list.
# Returns one row per table with step-level and row-level failure counts.
summarise_validation <- function(validation_list) {
  imap(validation_list, \(agent, table_name) {
    if (isTRUE(agent$.empty)) {
      return(tibble(
        table = table_name,
        n_steps = 0L,
        n_steps_failed = 0L,
        rows_failed = 0L,
        pct_steps_pass = 0
      ))
    }
    agent$validation_set |>
      summarise(
        table = table_name,
        n_steps = n(),
        n_steps_failed = sum(n_failed > 0, na.rm = TRUE),
        rows_failed = sum(n_failed, na.rm = TRUE),
        pct_steps_pass = mean(all_passed, na.rm = TRUE)
      )
  }) |>
    list_rbind()
}

# Run validation across all successful calls and return a long tidy tibble:
# one row per (call × table), with pdf / models / tokens / replicate attached.
#
# For an overall pass rate per call:
#   validate_all_calls(calls_results) |>
#     group_by(pdf, models, tokens, replicate) |>
#     summarise(pct_pass = mean(pct_steps_pass))
validate_all_calls <- function(calls_results) {
  calls_results |>
    filter(map_lgl(call, \(c) isTRUE(c$success))) |>
    mutate(
      validation_summary = map(call, \(c) {
        validate_llm_extraction(c$result) |>
          summarise_validation()
      })
    ) |>
    select(id, pdf, models, tokens, replicate, validation_summary) |>
    unnest(validation_summary)
}
```


### Comparison with Canonical Data

To determine to what extent the LLM-extracted data agreed with human-extracted data, calls were compared to "canonical" datasets extracted using the eData application using LLMs and human processing. Not all extracted data fields are directly comparable, but numerical and controlled vocabulary variables were compared (see @tbl-canonical-comparison-variables). Compared variables were converted to strings and checked for strict equality. In tables consisting of multiple rows of data (e.g. sites, parameters), we checked only that LLM-returned data existed in the canonical dataset.

Proportion of correct data was calculated relative to totals, and a score between 0 and 1 was returned to indicate degree of agreement.

```{r}
#| label: canon-comparison-setup
# --- Variable classification --------------------------------------------------
# var_type:
#   "controlled" — compare with case-insensitive exact match, return boolean
#   "continuous" — compute numeric delta (LLM − canonical)
#   "exclude"  — excluded from comparison
#
# canonical_name uses lowercase column names (matching rename_with(tolower) above).
# LLM schema names that differ from canonical names are mapped explicitly
# (e.g. country_iso -> country, ocean_iho -> area).

variable_meta <- tribble(
  ~table         , ~llm_name                     , ~canonical_name               , ~var_type    ,
  # campaign
  "campaign"     , "campaign_start_date"         , "campaign_start_date"         , "continuous" ,
  "campaign"     , "campaign_end_date"           , "campaign_end_date"           , "continuous" ,
  "campaign"     , "campaign_name"               , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_name_short"         , NA_character_                 , "exclude"    ,
  "campaign"     , "organisation"                , "organisation"                , "exclude"    ,
  "campaign"     , "campaign_comment"            , NA_character_                 , "exclude"    ,
  "campaign"     , "campaign_rationale"          , NA_character_                 , "exclude"    ,
  # references
  "references"   , "reference_type"              , "reference_type"              , "controlled" ,
  "references"   , "year"                        , "year"                        , "continuous" ,
  "references"   , "volume"                      , "volume"                      , "continuous" ,
  "references"   , "issue"                       , "issue"                       , "continuous" ,
  "references"   , "author"                      , NA_character_                 , "exclude"    ,
  "references"   , "title"                       , NA_character_                 , "exclude"    ,
  "references"   , "doi"                         , NA_character_                 , "exclude"    ,
  "references"   , "periodical_journal"          , NA_character_                 , "exclude"    ,
  "references"   , "publisher"                   , NA_character_                 , "exclude"    ,
  # sites
  "sites"        , "latitude"                    , "latitude"                    , "continuous" ,
  "sites"        , "longitude"                   , "longitude"                   , "continuous" ,
  "sites"        , "country_iso"                 , "country"                     , "controlled" ,
  "sites"        , "ocean_iho"                   , "area"                        , "controlled" ,
  "sites"        , "site_geographic_feature"     , "site_geographic_feature"     , "controlled" ,
  "sites"        , "site_geographic_feature_sub" , "site_geographic_feature_sub" , "controlled" ,
  "sites"        , "site_code"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_name"                   , NA_character_                 , "exclude"    ,
  "sites"        , "site_comment"                , NA_character_                 , "exclude"    ,
  # parameters
  "parameters"   , "parameter_type"              , "parameter_type"              , "controlled" ,
  "parameters"   , "cas_rn"                      , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_name"              , NA_character_                 , "exclude"    ,
  "parameters"   , "parameter_comment"           , NA_character_                 , "exclude"    ,
  # compartments
  "compartments" , "environ_compartment"         , "environ_compartment"         , "controlled" ,
  "compartments" , "environ_compartment_sub"     , "environ_compartment_sub"     , "controlled" ,
  "compartments" , "measured_category"           , "measured_category"           , "controlled" ,
  # biota
  "biota"        , "species_group"               , "species_group"               , "controlled" ,
  "biota"        , "sample_tissue"               , "sample_tissue"               , "controlled" ,
  "biota"        , "sample_species_lifestage"    , "sample_species_lifestage"    , "controlled" ,
  "biota"        , "sample_species_gender"       , "sample_species_gender"       , "controlled" ,
  "biota"        , "sample_species"              , NA_character_                 , "exclude"    ,
  "biota"        , "sample_id"                   , NA_character_                 , "exclude"    ,
  # methods
  "methods"      , "protocol_category"           , "protocol_category"           , "controlled" ,
  "methods"      , "protocol_name"               , "protocol_name"               , "controlled" ,
  "methods"      , "protocol_comment"            , NA_character_                 , "exclude"    ,
)

table_keys <- list(
  campaign = NULL,
  references = NULL,
  sites = c("latitude", "longitude"),
  parameters = "parameter_name",
  compartments = c("environ_compartment", "environ_compartment_sub"),
  biota = c("sample_species", "sample_tissue"),
  methods = c("protocol_category", "protocol_name")
)
```

[@tbl-canonical-comparison-variables] we check controlled and numerical variables against the "canonical versions" from the source data. This isn't very sophisticated:
- we just check if strings match
- for tables (sites, parameters, etc.) with more than one row we only check that what's in the call table is also in the canonical table (so if the call over-matches we don't flag that as an issue)

# Results

## Call Status

[@tbl-batch-report] reports the details and status of each call. Broadly speaking, two types of errors are possible. In case of connection issue, system downtime, or a lack of API credit, the call will return an HTTP error (e.g. HTTP 400), and no data. However, it has also been observed that in some cases, the API response will be flagged "successful" by the external provider, but returned only empty data. This is believed to be the result of the extraction process hitting the token limit before completing its response; for example when an unusually large PDF is send with a too-small token limit. This sharp distinction between performance at different token limits 

::: {.callout-note collapse="true"}
## Per-Replicate Call Status Raw Data

```{r}
#| label: tbl-batch-report
#| tbl-cap: "Per-replicate API call status and error messages."

# --- Batch run status report -------------------------------------------------
# Inspect the call list-column directly — before any flattening — to report
# on successes and errors with token counts, broken down by pdf, model, and
# token limit.

batch_run_report <- calls_results |>
  mutate(
    success = map_lgl(call, \(c) isTRUE(c$success)),
    error_msg = map_chr(call, \(c) c$error %||% NA_character_),
    # metadata is a named list of summed token counts for successful calls.
    # Try both naming conventions used across ellmer versions.
    input_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["input_tokens"]] %||% c$metadata[["input"]] %||% NA_real_
        ),
        error = \(e) NA_integer_
      )
    }),
    output_tokens = map_int(call, \(c) {
      tryCatch(
        as.integer(
          c$metadata[["output_tokens"]] %||%
            c$metadata[["output"]] %||%
            NA_real_
        ),
        error = \(e) NA_integer_
      )
    })
  )

# Per-replicate detail
batch_run_report |>
  select(
    id,
    pdf,
    models,
    tokens,
    replicate,
    success,
    error_msg
  ) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    id = "Call ID",
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    replicate = "Rep",
    success = "Success",
    error_msg = "Error"
  ) |>
  color(i = ~ !success, j = ~ success + error_msg, color = "firebrick")
```

:::

```{r}
#| label: tbl-batch-report-summary
#| tbl-cap: "Summary of API call success rates and mean token usage by PDF, model, and token limit. Note: Only contains data on most recent API call, *not* all cached calls."

# Grouped summary: counts and mean token usage by pdf × model × token limit
batch_run_report_summary <- batch_run_report |>
  group_by(pdf, models, tokens) |>
  summarise(
    id = paste0(id, collapse = ", "),
    n_calls = n(),
    n_success = sum(success),
    n_failed = sum(!success),
    error_types = paste(sort(unique(na.omit(error_msg))), collapse = "; "),
    mean_input_tok = round(mean(input_tokens, na.rm = TRUE)),
    mean_output_tok = round(mean(output_tokens, na.rm = TRUE)),
    .groups = "drop"
  )

batch_run_report_summary |>
  flextable() |>
  autofit() |>
  set_header_labels(
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    n_calls = "N",
    n_success = "Success",
    n_failed = "Failed",
    error_types = "Error Types",
    mean_input_tok = "Mean Input Tok",
    mean_output_tok = "Mean Output Tok"
  )
```

TODO: Some of these input tokens are rather suspect. Is stuff getting cached? Should we use the max instead?

::: {.callout-note collapse="true"}
## All Calls Raw Data

```{r}
#| label: tbl-all-calls
#| tbl-cap: "Status of all API calls, including both cached and recent calls."

call_status |> flextable() |> autofit()
```

:::


```{r}
#| label: tbl-all-calls-summary
#| tbl-cap: "Call completeness summary by PDF, model, and token limit."

call_status |>
  group_by(pdf, models, tokens) |>
  summarise(
    IDs = paste0(id, collapse = ", "),
    n_calls = n(),
    n_complete = sum(effectively_complete),
    n_empty_tables = sum(api_success & !effectively_complete),
    n_failed = sum(!api_success),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit()
```

```{r}
#| label: tbl-all-calls-summary-by-model
#| tbl-cap: "Failure rate by model"

model_failure_rate <- call_status |>
  group_by(models, tokens) |>
  summarise(
    n_calls = n(),
    n_complete = sum(effectively_complete),
    n_empty_tables = sum(api_success & !effectively_complete),
    n_failed = sum(!api_success),
    prop_failed = (n_failed + n_empty_tables) / n_calls,
    .groups = "drop"
  )

model_failure_rate |>
  flextable() |>
  set_header_labels(
    models = "Model",
    tokens = "Tokens",
    n_calls = "Calls",
    n_complete = "Complete",
    n_empty_tables = "Returned Empty",
    n_failed = "Call Failed",
    prop_failed = "Failure Rate"
  ) |>
  colformat_double(j = "prop_failed", digits = 3) |>
  autofit()
```

## eData Validation

```{r}
#| label: validation-execute

# Keep named agents available for get_agent_report() inspection.
# Index by row-number ID from batch_run_report, e.g. call_validations[["42"]]

# Reuse batch_run_report to identify which calls have a result worth validating.
successful_ids <- batch_run_report$id[batch_run_report$success]
call_validations <- calls[successful_ids] |>
  set_names(successful_ids) |>
  map(\(c) validate_llm_extraction(c$result))

table_summary <- validate_all_calls(calls_results)
```


```{r}
#| label: fig-validation-scores
#| fig-cap: "Proportion of validation steps passed by table and model. Each dot represents one replicate; dots stack at exact values to reflect the discrete nature of step-pass proportions."

table_summary |>
  mutate(tokens = factor(tokens)) |>
  ggplot(aes(x = table, y = pct_steps_pass, fill = tokens)) +
  geom_dotplot(
    binaxis = "y",
    stackdir = "center",
    binwidth = 0.04,
    dotsize = 0.5,
    position = position_dodge(width = 0.5)
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    name = "Proportion of Validation Passed"
  ) +
  facet_wrap(vars(models))
```


```{r}
#| label: fig-validation-relationship
#| fig-cap: "Across all tables, is there a relationship between token count and validation? Trend lines based on mean at each token level."

table_summary_provider <- table_summary |>
  left_join(models_tbl, by = c(models = "model_id")) |>
  select(
    id,
    pdf,
    models,
    tokens,
    replicate,
    table,
    n_steps,
    n_steps_failed,
    rows_failed,
    pct_steps_pass,
    provider
  )

make_provider_plot <- function(data, prov, palette, title) {
  d <- data |> filter(provider == prov)

  ggplot(d, aes(x = tokens, y = pct_steps_pass, colour = models)) +
    geom_point(position = position_jitter(width = 500, height = 0), shape = 1) +
    stat_summary(
      aes(x = tokens, colour = models, group = models),
      fun = mean,
      geom = "line"
    ) +
    scale_colour_viridis_d(option = palette, name = NULL) +
    scale_y_continuous(
      limits = c(0, 1),
      name = "Proportion of Validation Passed"
    ) +
    scale_x_continuous(limits = c(2000, 10000), breaks = c(3000, 6000, 9000)) +
    labs(title = title, x = "Max Tokens") +
    theme_minimal()
}

p_anthropic <- make_provider_plot(
  table_summary_provider,
  "Anthropic",
  "viridis",
  "Anthropic"
)
p_gemini <- make_provider_plot(
  table_summary_provider,
  "Google Gemini",
  "plasma",
  "Google Gemini"
)
p_openai <- make_provider_plot(
  table_summary_provider,
  "OpenAI",
  "cividis",
  "OpenAI"
)

p_anthropic / p_gemini / p_openai + plot_layout(axes = "collect")
```

@fig-validation-relationship describes the relationship between token count and proportion of validation passed, per provider and model. We see:

- Claude models show a clear pattern of frequent failures at 3000 tokens. At 6000 and 9000 tokens, success rates are roughly consistent
- Gemini-3.5-flash fails at all but 9000 tokens
- OpenAI's models have a consistent validation rate across models and token counts. 

```{r}
#| label: canonical-data-setup

pdf_zip_map <- c(
  a = here("data-raw/1997RitterhoffTraceMetalsIn_AllData_20251030_145522.zip"),
  b = here(
    "data-raw/2002HellandTransportAndSedimentation_AllData_20251104_075533.zip"
  ),
  c = here(
    "data-raw/2021WojtasikTemporalIsolationBetween_AllData_20251030_163338.zip"
  ),
  d = here(
    "data-raw/2022KohlerDistributionPatternOf_AllData_20251217_115443.zip"
  )
)

load_canonical <- function(zip_path) {
  tmp <- tempfile()
  dir.create(tmp)
  unzip(zip_path, exdir = tmp)
  csvs <- list.files(tmp, full.names = TRUE, pattern = "\\.csv$")

  patterns <- c(
    campaign = "Campaign",
    references = "Reference",
    sites = "Sites",
    parameters = "Parameters",
    compartments = "Compartments",
    biota = "Biota",
    methods = "Methods"
  )

  imap(patterns, function(pat, nm) {
    f <- csvs[grepl(pat, csvs, fixed = TRUE)]
    if (length(f) == 0) {
      return(NULL)
    }
    read_csv(f[1], show_col_types = FALSE) |> rename_with(tolower)
  }) |>
    compact()
}

# Load canonical data for all four papers (cached in-memory)
canonical_data <- imap(pdf_zip_map, \(path, id) load_canonical(path))
```




::: {.callout-note collapse="true"}
## Canon Comparison Raw Data

```{r}
#| label: canon-comparison-all

# Scale string comparison across all calls × all tables.
#
# For each (call_id, table) pair:
#   1. Align LLM field names to canonical field names via variable_meta.
#   2. Pool all values across rows (ignoring row identity).
#   3. Per field: any LLM value == any canonical value → match.
# Produces one row per (call × table) with n_match, n_total, prop_match.

# Step 1: Flatten canonical_data (pdf → table → data.frame) into a single long
# data frame of (pdf, table, canonical_name, value_canon)
canonical_long <- list()
for (pdf_id in names(canonical_data)) {
  for (tbl in names(canonical_data[[pdf_id]])) {
    tbl_df <- canonical_data[[pdf_id]][[tbl]]
    if (!is.null(tbl_df) && nrow(tbl_df) > 0) {
      key <- paste(pdf_id, tbl, sep = "__")
      canonical_long[[key]] <- tbl_df |>
        pivot_longer(cols = everything(), values_transform = as.character) |>
        mutate(pdf = pdf_id, table = tbl)
    }
  }
}
canonical_long <- bind_rows(canonical_long) |>
  rename(canonical_name = name, value_canon = value)

# Step 2: Filter variable_meta once to reuse below
meta_filtered <- variable_meta |>
  filter(!is.na(var_type), var_type != "exclude")

# Step 3: Build the base grid of every (call, table) combination to evaluate
call_table_combos <- calls_flat |>
  distinct(id, pdf, models, tokens, replicate) |>
  crossing(table = meta_filtered |> distinct(table) |> pull(table))

# Step 4: LLM values — pull from calls_flat and map field names to canonical names
llm_vals_all <- calls_flat |>
  inner_join(
    meta_filtered |> select(table, llm_name, canonical_name),
    by = c(level_2 = "table", name = "llm_name")
  ) |>
  select(id, table = level_2, canonical_name, value_call = value)

# Step 5: Canonical values — keep only fields present in meta
canon_vals_all <- canonical_long |>
  inner_join(
    meta_filtered |> select(table, canonical_name),
    by = c("table", "canonical_name")
  ) |>
  select(pdf, table, canonical_name, value_canon)

# Step 6: Join and compute match statistics
# - semi_join: restrict to pdf+table combos that have any canonical data
# - inner_join llm_vals: restrict to combos that have any LLM output for this table
# - left_join canon_vals: keep all LLM fields; unmatched canon fields → match = FALSE
canonical_comparison <- call_table_combos |>
  semi_join(canon_vals_all, by = c("pdf", "table")) |>
  inner_join(llm_vals_all, by = c("id", "table")) |>
  left_join(
    canon_vals_all,
    by = c("pdf", "table", "canonical_name"),
    relationship = "many-to-many"
  ) |>
  # Per field: does any LLM value match any canonical value?
  group_by(id, pdf, models, tokens, replicate, table, canonical_name) |>
  summarise(
    match = any(value_call == value_canon, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # Per call+table: aggregate match statistics
  group_by(id, pdf, models, tokens, replicate, table) |>
  summarise(
    n_match = sum(match, na.rm = TRUE),
    n_total = n(),
    prop_match = n_match / n_total,
    .groups = "drop"
  )

canonical_comparison |>
  mutate(id = as.integer(id)) |>
  arrange(id) |>
  flextable() |>
  autofit() |>
  colformat_double(j = "prop_match", digits = 2) |>
  set_header_labels(
    id = "Call ID",
    pdf = "PDF",
    models = "Model",
    tokens = "Token Limit",
    replicate = "Rep",
    table = "Table",
    n_match = "N Matched",
    n_total = "N Total",
    prop_match = "Prop. Matched"
  )
```

:::

Our smaller token calls (3000) end up all excluded here, because with all NA data there's nothing to compare.

```{r}
#| label: fig-canon-comparison-summary
#| fig-cap:
#|   - "Mean proportion of fields matching canonical data, by model and token limit."
#|   - "Proportion of fields matching canonical data, by model and table."
# what's the best way to summarise this?
canonical_comparison_summarised <- canonical_comparison |>
  group_by(models, tokens) |>
  replace_na(list(prop_match = 0)) |>
  reframe(mean_prop_match = mean(prop_match), n_replicates = max(replicate)) |>
  arrange(desc(mean_prop_match))

# our questions are:
# which models are most accurate?
# which tables cause the most problems?
# does token count affect anything?

# for a standard "look how good my LLM is" comparison, we'd just do a bar chart...

canonical_comparison |>
  # recode NA as 0 so we don't get unfairly high averages
  replace_na(list(prop_match = 0)) |>
  # summarise by call (dropping table) to reduce noisyness of points
  reframe(
    .by = id,
    mean_prop_match = mean(prop_match),
    tokens,
    models,
    pdf,
    replicate
  ) |>
  ggplot(aes(
    x = ordered(tokens),
    y = mean_prop_match,
    fill = ordered(tokens),
    group = ordered(tokens)
  )) +
  geom_point(
    position = position_jitter(width = 0.3, height = 0),
    aes(shape = pdf)
  ) +
  geom_boxplot(alpha = 0.75) +
  scale_colour_viridis_d(
    breaks = c("3000", "6000", "9000"),
    name = "Max tokens"
  ) +
  scale_fill_viridis_d(
    breaks = c("3000", "6000", "9000"),
    name = "Max tokens"
  ) +
  scale_y_continuous(
    # limits = c(0, 1),
    name = "Proportion of data matching canonical"
  ) +
  scale_x_discrete(breaks = c("3000", "6000", "9000"), name = "Max tokens") +
  geom_text(
    data = model_failure_rate |>
      mutate(
        tokens = ordered(tokens),
        label = scales::percent(prop_failed, accuracy = 1)
      ),
    aes(
      x = tokens,
      y = Inf,
      label = label,
      colour = if_else(prop_failed > 0.1, "red", "grey40")
    ),
    inherit.aes = FALSE,
    vjust = 1.5,
    size = 2.8
  ) +
  scale_colour_identity() +
  facet_wrap(facets = vars(models), nrow = 1) +
  theme_minimal()
```


## Cost Analysis

```{r}
#| label: cost-analysis

# Parse the cost string columns from models_tbl into numeric $/Mtok values.
# Handles:
#   "$1"                              → 1
#   "1.25 (<= 200k tokens), then 2.5" → 1.25  (use the lower-tier price)
#   "Free"                            → 0
#   placeholder strings (no digits)   → NA
parse_cost <- function(x) {
  # Extract the first number in the string; return 0 for "Free", NA otherwise
  case_when(
    str_to_lower(x) == "free" ~ 0,
    str_detect(x, "[0-9]") ~ as.numeric(str_extract(x, "[0-9]+\\.?[0-9]*")),
    .default = NA_real_
  )
}

model_costs <- models_tbl |>
  select(model_id, cost_input_mtok, cost_output_mtok) |>
  mutate(
    cost_input_per_mtok = parse_cost(cost_input_mtok),
    cost_output_per_mtok = parse_cost(cost_output_mtok)
  )

# Join token counts from batch_run_report with per-model pricing,
# then compute estimated USD cost for each call.
call_costs <- batch_run_report |>
  left_join(model_costs, by = c("models" = "model_id")) |>
  mutate(
    cost_usd = (input_tokens / 1e6) *
      cost_input_per_mtok +
      (output_tokens / 1e6) * cost_output_per_mtok
  ) |>
  select(-model_provider) # TODO: Why in there twice?
```

Costs are necessarily an estimate and rather rough - most models provide a cost per token value, and most calls return a token count... but not all of them are trustworthy. I suspect that behind the scenes some data is getting cached between calls, which means subsequent calls with the same/similar data are much cheaper. This is probably a good thing, but I'm not sure how to deal with this in the analysis.

```{r}
#| label: tbl-cost-summary
#| tbl-cap: "Estimated API cost per call (USD) by model and token limit. Costs are computed from actual input and output token counts reported by the API, multiplied by the published per-million-token rates."

call_costs |>
  group_by(models, tokens) |>
  summarise(
    n_calls = n(),
    n_success = sum(success),
    mean_input_tok = round(mean(input_tokens, na.rm = TRUE)),
    mean_output_tok = round(mean(output_tokens, na.rm = TRUE)),
    mean_cost_usd = mean(cost_usd, na.rm = TRUE),
    total_cost_usd = sum(cost_usd, na.rm = TRUE),
    .groups = "drop"
  ) |>
  flextable() |>
  autofit() |>
  colformat_double(j = c("mean_cost_usd", "total_cost_usd"), digits = 5) |>
  set_header_labels(
    models = "Model",
    tokens = "Token Limit",
    n_calls = "N Calls",
    n_success = "N Success",
    mean_input_tok = "Mean Input Tok",
    mean_output_tok = "Mean Output Tok",
    mean_cost_usd = "Mean Cost (USD)",
    total_cost_usd = "Total Cost (USD)"
  )
```

Surprisingly, cost and max token count correlate for Anthropic models, but don't seem to for OpenAI [@fig-cost-by-model-tokens]. This is a surprise. Note the drop in mean costs for Anthropic (and in some cases for OpenAI) at 6000 tokens. I'm fairly sure this is a consequence of caching (as discussed above), while the 9000-token calls were run a day later. I can try and turn caching off and re-run for more accurate numbers, though this may be expensive and time-consuming. 

```{r}
#| label: fig-cost-by-model-tokens
#| fig-cap: "Estimated cost per call (USD) by model and token limit, split by provider. Each point is one replicate call; lines connect group means."

make_cost_plot <- function(data, prov, palette, title) {
  d <- data |> filter(provider == prov)
  ggplot(d, aes(x = tokens, y = cost_usd, colour = models)) +
    geom_point(
      position = position_jitter(width = 500, height = 0),
      shape = 1,
      alpha = 0.6
    ) +
    stat_summary(aes(group = models), fun = mean, geom = "line") +
    scale_colour_viridis_d(option = palette, name = NULL) +
    scale_x_continuous(limits = c(2000, 10000), breaks = c(3000, 6000, 9000)) +
    scale_y_continuous(name = "Estimated Cost (USD)", limits = c(0, 0.3)) +
    labs(title = title, x = "Max Tokens") +
    theme_minimal()
}

pc_anthropic <- make_cost_plot(
  call_costs,
  "Anthropic",
  "viridis",
  "Anthropic"
)
pc_gemini <- make_cost_plot(
  call_costs,
  "Google Gemini",
  "plasma",
  "Google Gemini"
)
pc_openai <- make_cost_plot(call_costs, "OpenAI", "cividis", "OpenAI")

pc_anthropic / pc_gemini / pc_openai + plot_layout(axes = "collect")
```

```{r}
#| label: fig-cost-vs-quality
#| fig-cap: "Cost vs. extraction quality (mean proportion of canonical fields matched), by model and token limit. Each point is one replicate; a higher proportion matched at lower cost is preferable."

cost_quality <- call_costs |>
  filter(success) |>
  select(id, pdf, models, tokens, replicate, cost_usd) |>
  mutate(id = as.character(id)) |>
  left_join(
    canonical_comparison_summarised,
    by = c("models", "tokens")
  ) |>
  replace_na(list(mean_prop_match = 0, n_replicates = 3))

cost_quality |>
  mutate(tokens = factor(tokens)) |>
  ggplot(aes(
    x = cost_usd,
    y = mean_prop_match,
    colour = models,
    shape = tokens
  )) +
  geom_point(size = 3, alpha = 0.8) +
  # scale_y_continuous(
  #   limits = c(0, 1),
  #   name = "Mean Proportion of Fields Matched"
  # ) +
  labs(
    x = "Estimated Cost (USD)",
    colour = "Model",
    shape = "Token Limit"
  )
```

```{r}
#| label: fig-cost-vs-validation
#| fig-cap: "Estimated cost per call (USD) vs. mean eData validation rate (proportion of validation steps passed), split by provider. Each point is one replicate; higher validation at lower cost is preferable. Failed calls (i.e. mean validation pass rate of 0) excluded."

cost_validation <- call_costs |>
  filter(success) |>
  select(id, models, tokens, replicate, cost_usd, provider) |>
  left_join(
    table_summary |>
      group_by(id, models, tokens, replicate) |>
      summarise(
        mean_pct_pass = mean(pct_steps_pass, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("id", "models", "tokens", "replicate")
  ) |>
  filter(!is.na(mean_pct_pass), mean_pct_pass != 0)

make_cost_validation_plot <- function(data, prov, palette, title) {
  d <- data |> filter(provider == prov)
  ggplot(d, aes(x = cost_usd, y = mean_pct_pass, colour = models)) +
    geom_point(size = 2.5, alpha = 0.7) +
    scale_colour_viridis_d(option = palette, name = NULL) +
    scale_y_continuous(limits = c(0, 1), name = "Mean Validation Rate") +
    scale_x_continuous(limits = c(0, 0.3)) +
    labs(title = title, x = "Estimated Cost (USD)") +
    theme_minimal()
}

pcv_anthropic <- make_cost_validation_plot(
  cost_validation,
  "Anthropic",
  "viridis",
  "Anthropic"
)
pcv_gemini <- make_cost_validation_plot(
  cost_validation,
  "Google Gemini",
  "plasma",
  "Google Gemini"
)
pcv_openai <- make_cost_validation_plot(
  cost_validation,
  "OpenAI",
  "cividis",
  "OpenAI"
)

pcv_anthropic / pcv_gemini / pcv_openai + plot_layout(axes = "collect")
```


## Results Summary Table

@tbl-results-summary still isn't very illuminating. 

```{r}
#| label: tbl-results-summary
#| tbl-cap: "Summary of validation accuracy (mean score, 1–5 scale), canonical field matching (mean proportion of canonical fields recovered), and estimated cost per call (USD), by model and token limit. Each row aggregates across all papers and replicates evaluated."
#| echo: false

# Validation scores: mean score per model × tokens (excluding NAs)
val_summary <- scores_long |>
  filter(!is.na(score)) |>
  group_by(models, tokens) |>
  summarise(
    n_papers = n_distinct(pdf),
    n_replicates = n_distinct(replicate),
    mean_val_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  )

# Canonical matching: mean proportion of fields matched per model × tokens
canon_summary <- canonical_comparison |>
  group_by(models, tokens) |>
  summarise(
    mean_prop_match = mean(prop_match, na.rm = TRUE),
    .groups = "drop"
  )

# Cost: mean USD per call per model × tokens
cost_summary <- call_costs |>
  filter(success) |>
  group_by(models, tokens) |>
  summarise(
    mean_cost_usd = mean(cost_usd, na.rm = TRUE),
    .groups = "drop"
  )

# Combine
val_summary |>
  left_join(canon_summary, by = c("models", "tokens")) |>
  left_join(cost_summary, by = c("models", "tokens")) |>
  arrange(models, tokens) |>
  flextable() |>
  autofit() |>
  colformat_double(j = "mean_val_score", digits = 2) |>
  colformat_double(j = "mean_prop_match", digits = 3) |>
  colformat_double(j = "mean_cost_usd", digits = 5) |>
  set_header_labels(
    models = "Model",
    tokens = "Token Limit",
    n_papers = "Papers",
    n_replicates = "Replicates",
    mean_val_score = "Mean Validation Score",
    mean_prop_match = "Mean Canonical Match",
    mean_cost_usd = "Mean Cost (USD)"
  )
```

# Appendices

```{r}
#| label: tbl-canonical-comparison-variables
#| tbl-cap: "List of tables and variables returned by LLM extraction, alongside whether these variables are used in the comparison (controlled vocabulary or numeric values), or are skipped (free or descriptive text)."
#| echo: false

variable_meta |>
  mutate(
    llm_name = toupper(llm_name),
    table = str_to_sentence(table),
    var_type = case_when(
      var_type %in% c("controlled", "continuous") ~ "Compared",
      var_type == "exclude" ~ "Skipped",
      .default = str_to_sentence(var_type)
    )
  ) |>
  select(table, llm_name, var_type) |>
  flextable() |>
  autofit() |>
  set_header_labels(
    table = "Table",
    llm_name = "Variable Name",
    var_type = "Variable Check"
  )
```