---
title: "LLM Evaluation"
subtitle: "Comparing the quality, speed, and cost of LLM extraction"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
library(STOPeData)
library(ellmer)
library(tidyverse)
library(flextable)
library(arrow)

devtools::load_all()
```

The `ellmer` package includes functions that connect to a wide variety of local and remote APIs from a variety of providers. Put simply, we have a great many options. Which are better, and which are worse? We might assume that more expensive models are more effective at data extraction, and that larger and more complex files are more difficult - and require more tokens - than simpler models when extracting data. But can we prove any of these assumptions?

The question of how we evaluate different LLMs is an important one. I'm not quite sure how to do it. There are some obvious measures:

- Time
- Cost
- Does the model fail entirely?
- Does the model hit a token limit, or otherwise fail totally?
- Is the data returned obviously wrong/incomplete?

All of these are easy to measure, but the question of the overall quality and accuracy of extracted data are more complicated. We can run our `eDataDRF` validation functions against whatever we get out, but they won't be a perfect fit as they're designed to work with data after human extraction. Still, it may be the easiest approach.

We definitely want to evaluate our "recommended" models from Anthropic, OpenAI and Google. I don't anticipate any major differences. I'd also like to try some of the most popular models from LM Studio, as it supports PDF data extraction out the box, although I haven't been able to get it working yet. 

I then suggest some sort of the following experimental design:

- 3 providers x 2 models, + 1 LM Studio (n = 7)
- 3 different papers of varying lengths (n = 3)
- 3 different token limits, small, medium, and large (n = 3)
- 3 replicates (n = 3) # increase once we've tested it

This will give us 7 * 27 = 189 cases. Which is quite a lot, but we can probably batch them and let them get on with themselves. 

Use multiple models to do a consensus thing and maybe compare against people doing the data extraction manually

```{r}
#| label: pdfs

pdfs <- tibble(
  id = c("a", "b", "c"),
  path = c(
    "data-raw/norheim-1987.pdf",
    "data-raw/wojtasik-2021.pdf",
    "data-raw/zauke-1999.pdf"
  ),
  comment = c(
    "Well-OCR'd text, some tables (difficult to read). Copper concentrations reported explicitly in tables. Score: 4.",
    "More recent, clear layout with some (irrelevant) images. Score: 4.",
    "Well laid out with extremely large quantity of data on different species."
  ),
  # Fill in after reviewing each PDF
  num_images = c(NA_integer_, NA_integer_, NA_integer_),
  num_tables = c(NA_integer_, NA_integer_, NA_integer_)
) |>
  mutate(
    info = map(path, pdftools::pdf_info),
    file_size_kb = map_dbl(path, \(p) file.size(p) / 1024),
    num_pages = map_int(info, \(x) x$pages)
  ) |>
  select(-info)

# TODO: Add more paper details, default rounding
pdfs |>
  flextable() |>
  autofit()
```

Sources for model comparisons: https://platform.claude.com/docs/en/about-claude/models/overview, https://developers.openai.com/api/docs/models/compare, https://ai.google.dev/gemini-api/docs/models


```{r}
#| label: models

models_tbl <- tribble(
  ~provider          , ~model_label        , ~model_id                ,
  ~release_date      , ~context_window_k   , ~max_output_k            ,
  ~cost_input_mtok   , ~cost_output_mtok   ,
  ~knowledge_cutoff  , ~training_cutoff    , ~niche                   ,
  # ---- Anthropic ---------------------------------------------------------------
  "Anthropic"        , "Claude Haiku 4.5"  , "claude-haiku-4-5"       ,
  "release_date"     , "200k"              , "64k"                    ,
  "$1"               , "$5"                ,
  "Feb 2025"         , "Jul 2025"          , "niche"                  ,
  #
  "Anthropic"        , "Claude Sonnet 4.6" , "claude-sonnet-4-6"      ,
  "release_date"     , "1M"                , "128k"                   ,
  "$3"               , "$15"               ,
  "Aug 2025"         , "Jan 2026"          , "niche"                  ,
  #
  "Anthropic"        , "Claude Opus 4.7"   , "claude-opus-4-7"        ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Anthropic"        , "Claude Opus 4.8"   , "claude-opus-4-8"        ,
  "release_date"     , "1M"                , "128k"                   ,
  "$5"               , "$25"               ,
  "Jan 2026"         , "Jan 2026"          , "niche"                  ,
  # ---- OpenAI ------------------------------------------------------------------
  # TODO: OpenAI distinguishes between input and cached input cost. Wwhat does this mean? I used regular input for now.
  # Also only provides "knowledge cutoff"
  "OpenAI"           , "GPT-5.4 Mini"      , "gpt-5.4-mini"           ,
  "release_date"     , "400"               , "128"                    ,
  "0.75"             , "4.50"              ,
  "Aug 2025"         , "training_cutoff"   , "niche"                  ,
  #
  "OpenAI"           , "GPT-5.4"           , "gpt-5.4"                ,
  "release_date"     , "1050"              , "128"                    ,
  "2.5"              , "15.0"              ,
  "Aug 2025"         , "training_cutoff"   , "niche"                  ,
  #
  "OpenAI"           , "GPT-5.5"           , "gpt-5.5"                ,
  "release_date"     , "1050"              , "128"                    ,
  "cost_input_mtok"  , "30.0"              ,
  "Dec 2025"         , "training_cutoff"   , "niche"                  ,
  # ---- Google Gemini -----------------------------------------------------------
  "Google Gemini"    , "Gemini 2.5 Pro"    , "gemini-2.5-pro"         ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Google Gemini"    , "Gemini 3.1 Pro"    , "gemini-3.1-pro-preview" ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"                  ,
  #
  "Google Gemini"    , "Gemini 3.5 Flash"  , "gemini-3.5-flash"       ,
  "release_date"     , "context_window_k"  , "max_output_k"           ,
  "cost_input_mtok"  , "cost_output_mtok"  ,
  "knowledge_cutoff" , "training_cutoff"   , "niche"
)
# Named vector for use in experimental_setup (label -> API id)
models <- setNames(models_tbl$model_id, models_tbl$model_label)


models_tbl |> flextable() |> autofit()
```

So, what experiments do we want to run?

```{r}
#| label: experimental_design

experimental_setup <-
  expand.grid(
    pdf = c("a", "b", "c"),
    models = models,
    tokens = c("1000", "5000", "10000"),
    replicate = 1:3
  ) |>
  as_tibble()
```


```{r}
#| label: API_calls

pdf_1 <- "data-raw/norheim-1987.pdf"


# Recursively walk the list, collecting path-as-vector as we go.
# Each branch terminates at: (1) a data frame, (2) an atomic scalar.
flatten_to_rows <- function(x, path = character()) {
  if (is.data.frame(x)) {
    # --- TIBBLE BRANCH ---
    # Capture column types before pivoting (pivot coerces everything to one type)
    col_types <- sapply(x, \(col) paste(class(col), collapse = "/"))

    x |>
      mutate(.row_idx = row_number()) |>
      pivot_longer(
        cols = -.row_idx,
        names_to = "name",
        values_to = "value",
        values_transform = as.character # coerce all cells → chr for uniform value col
      ) |>
      mutate(
        level_path = list(path), # store the full path as a list-col
        row_idx = .row_idx,
        type = col_types[name] # look up original type by col name
      ) |>
      select(level_path, name, row_idx, value, type)
  } else if (is.list(x)) {
    # --- LIST BRANCH: recurse into each named element ---
    imap(x, \(item, nm) flatten_to_rows(item, c(path, nm))) |>
      list_rbind()
  } else {
    # --- SCALAR BRANCH ---
    # The last segment of path IS the name; everything before is the path.
    tibble(
      level_path = list(head(path, -1)),
      name = tail(path, 1),
      row_idx = NA_integer_,
      value = as.character(x),
      type = paste(class(x), collapse = "/")
    )
  }
}

# TODO: here()

if (!file.exists("data-raw/llm_eval_output_flat.parquet")) {
  {
    call1 <- extract_pdf_with_llm(
      pdf = pdf_1,
      model_provider = "Anthropic",
      model_name = "claude-haiku-4-5",
      env_var = "ANTHROPIC_API_KEY",
      chat_fn = "chat_anthropic",
      extraction_prompt = create_extraction_prompt(),
      extraction_schema = create_extraction_schema(),
      api_key = Sys.getenv("ANTHROPIC_API_KEY"),
      max_tokens = 6000
    )
  }

  call2 <- extract_pdf_with_llm(
    pdf = pdf_1,
    model_provider = "Anthropic",
    model_name = "claude-sonnet-4-6",
    env_var = "ANTHROPIC_API_KEY",
    chat_fn = "chat_anthropic",
    extraction_prompt = create_extraction_prompt(),
    extraction_schema = create_extraction_schema(),
    api_key = Sys.getenv("ANTHROPIC_API_KEY"),
    max_tokens = 6000
  )

  # Run it, then unpack the path list-col into discrete level_1 / level_2 columns
  call1_flat <- flatten_to_rows(call1) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)

  call2_flat <- flatten_to_rows(call2) |>
    mutate(
      level_1 = map_chr(level_path, \(p) {
        if (length(p) >= 1) p[[1]] else NA_character_
      }),
      level_2 = map_chr(level_path, \(p) {
        if (length(p) >= 2) p[[2]] else NA_character_
      })
    ) |>
    select(level_1, level_2, name, row_idx, value, type)

  calls_flat <- bind_rows(call1_flat, call2_flat, .id = "id")

  write_parquet(calls_flat, "data-raw/llm_eval_output_flat.parquet")
} else {
  message("LLM API call cached, skipping execution and loading from /data-raw")
  calls_flat <- read_parquet("data-raw/llm_eval_output_flat.parquet")
}
```


```{r}
#| label: call-comparison
calls_flat

# KET is right, pairwise might get messy quickly

# campaign and reference are easy.

# almost certainly there are better options than identical()
compare_campaign <- calls_flat |>
  filter(level_2 == "campaign") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# no issues with references
compare_references <- calls_flat |>
  filter(level_2 == "references") |>
  group_by(name) |>
  distinct(value) |>
  reframe(n = n()) |>
  filter(n > 1)

# sites...
# issue: different calls return variables in different orders!
# we can't automatically match sites to each other unless we assume that the lat/long are always equal (or, if we have a canonical point of comparison, I guess)
compare_sites <- calls_flat |>
  filter(level_2 == "sites") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# we can presumably compare distance with sf

library(sf)
# for example
compare_sites |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric))

# how do we check if each value has an equivalent?
# I guess we just group and ignore idx
# we may have to call st_distance for g
distances <- calls_flat |>
  filter(level_2 == "sites") |>
  filter(name %in% c("latitude", "longitude")) |>
  mutate(across(contains("call"), as.numeric)) |>
  group_by(id, row_idx) |>
  pivot_wider(names_from = name, values_from = value) |>
  st_as_sf(coords = c("latitude", "longitude"), na.fail = FALSE)

# pair sites and check (great circle) distance?

x <- distances[1:3, 6]
y <- distances[4:6, 6]

# I guess we make a distance matrix, which should have one value of approximately 0 and be... orthogonal?
# I believe these distances are in km, would be good to check
st_distance(x, y)

# parameters should be easier, they have more-or-less canonical names
compare_parameters <- calls_flat |>
  filter(level_2 == "parameters") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# compartments - easy
compare_compartments <- calls_flat |>
  filter(level_2 == "compartments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# biota
# common or linnean name is acceptable as long as they agree
compare_biota <- calls_flat |>
  filter(level_2 == "biota") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# methods
# again, pairing may be an issue here
compare_methods <- calls_flat |>
  filter(level_2 == "methods") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# samples
# might not do the evaluation on this one, since we don't strictly speaking use it
compare_samples <- calls_flat |>
  filter(level_2 == "samples") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call")

# comments are pleasantly easy
compare_comments <- calls_flat |>
  filter(level_2 == "comments") |>
  group_by(name) |>
  pivot_wider(names_from = id, names_prefix = "call") |>
  mutate(
    score_1 = str_extract(call1, "(?!Score\\: )[1-5]"),
    score_2 = str_extract(call2, "(?!Score\\: )[1-5]"),
    .before = call1
  )
```